\documentclass[main.tex]{subfiles}
\begin{document}

\section{Sheet 3}

\subsection{Riemann tensor computations}

\subsubsection{LIF form}

In the LIF, we have \(g_{\mu \nu } = \eta_{\mu \nu }\) and \(= g_{\mu \nu , \alpha }=0\). Therefore, the Christoffel symbols are all zero. 
In the explicit expression of the Riemann tensor, which looks like \(R = \partial \Gamma  + \Gamma \Gamma\) we can drop the second term and keep only the derivatives of the Christoffel symbols, which are nonzero since they depend on the second derivatives of the metric.

If the Christoffel symbols are zero, the computation of the curvature tensor is significantly easier: 
%
\begin{subequations}
\begin{align}
  R^{\mu }_{\nu \rho \sigma }
  &=  2 \partial_{[\rho }\Gamma^{\mu }_{\sigma ] \nu}  \\
  &= 2 \partial_{[\rho } \qty(\frac{1}{2} g^{\mu \alpha } \qty(g_{\alpha |\sigma ], \nu } + g_{\alpha \nu  , |\sigma  ]} - g_{\nu  |\sigma], \alpha }))  \\
  &= g^{\mu \alpha }\qty(g_{\alpha [\sigma |, \nu | \rho      ]} + \cancelto{}{g_{\alpha \nu, [\sigma \rho]}}-g_{\nu [\sigma |, \alpha | \rho  ]})
\,.
\end{align}
\end{subequations}

So, if we lower the index of the (1, 3) Riemann tensor we get the desired expression for the all-lower (0, 4) Riemann tensor: 
%
\begin{equation}
  R_{\mu \nu \rho \sigma} = g_{\mu [\sigma |, \nu |\rho ]} - g_{\nu  [\sigma |, \mu | \rho ]} 
\,.
\end{equation}

\subsubsection{Ricci tensor and scalar}

The Ricci tensor is given by: 
%
\begin{equation}
  R_{\mu \nu } = R^{\alpha }_{\mu \alpha \nu } 
  = g^{\alpha \beta }R_{\alpha \mu \beta \nu }
  = g^{\alpha \beta }\qty(g_{\alpha [\nu|, \mu | \beta ]} - g_{\mu [\nu|, \alpha | \beta ]})
\,.
\end{equation}
%
% where we can simplify all the antisymmetric terms in \(\alpha \beta \), since they are contracted with a symmetric tensor.

Now, of the four terms in the sum one simplifies immediately: we know that the Frobenius norm of the metric is constant: \(g^{\alpha \beta }g_{\alpha \beta } \equiv 4\), so if we expand we get \(0=\frac[i]{1}{2} (g^{\alpha \beta }g_{\alpha \beta })_{,\nu \mu } = g^{\alpha \beta }g_{\alpha \beta , \nu \mu } + g^{\alpha \beta }\,_{,\nu }g_{\alpha \beta , \mu }\), but the first derivatives of the metric vanish in the LIF so we only keep the first term, therefore \(g^{\alpha \beta }g_{\alpha \beta , \nu \mu }=0\).

So we have 
%
\begin{subequations}
    \begin{align}
    R_{\mu \nu }
    &= g^{\alpha \beta }\qty(
      g_{\alpha \nu , \mu \beta }
      -\cancelto{}{g_{\alpha \beta , \mu \nu }}
      - g_{\mu \nu , \alpha \beta }
      + g_{\mu \beta , \alpha \nu }
    )  \\
    &=g^{\alpha \beta }\qty(
      2 g_{\alpha (\nu , \mu ) \beta }
      - g_{\mu \nu , \alpha \beta }
    )  \\
    &= 2 g_{\alpha (\nu , \mu )}\,\!^{\alpha } - \square g_{\mu \nu } \label{eq:Ricci-tensor-LIF}
\,,
\end{align}
\end{subequations}
%
where\footnote{All the indices after the comma are derivatives, even when they become upper.} the square \(\square\) denotes the D'alambert operator, \(\square = \partial_{\beta }\partial^{\beta }\).

The Ricci scalar, on the other hand, is given by 
%
\begin{equation} \label{eq:ricci-scalar-LIF} 
  R = g^{\mu \nu }R_{\mu \nu }
  = 2 g^{\mu \nu }g_{\alpha (\nu , \mu )}\,\!^{\alpha } - \cancelto{}{g^{\mu \nu }\square g_{\mu \nu }}
  = 2 \tensor{g}{_{\alpha \nu }^{, \alpha \nu }}
\,,
\end{equation}
%
where the last term vanished since \(\square \qty(g^{\mu \nu }g_{\mu \nu })=0\), and while expanding we get twice \(g^{\mu \nu }\square g_{\mu \nu }\) since the first derivatives are zero.

We neglected the symmetrization in the first term since when contracting with the metric the indices \(\mu \nu \) are automatically symmetrized.

\subsubsection{LIF identities}

%
\begin{subequations}
\begin{align}
  \nabla_{\alpha }\qty(R^{ \alpha }_{\beta } - \frac{R}{2} \delta^{\alpha }_{\beta }) &=  \partial_{\alpha }\qty(g^{\alpha \lambda }R_{\lambda \beta } - \frac{R}{2} \delta^{\alpha }_{\beta })  \\
  &= \eta^{\alpha \lambda }R_{\lambda \beta, \alpha} - \frac{1}{2} R_{,\beta } \label{eq:contracted-Bianchi-LIF}
\,,
\end{align}
\end{subequations}
%
since in the LIF the metric is the Minkowski one, and the Christoffel symbols are zero therefore \(\nabla_{\alpha }= \partial_{\alpha }\).

\subsubsection{Contracted Bianchi identities}

The first term is
%
\begin{equation}
  \eta^{\beta \mu }R_{\mu \nu, \beta } = 
  \eta^{\beta \mu }\qty(2 \tensor{g}{_{\alpha (\nu , \mu )}^{,\alpha }} - \square g_{\mu \nu })_{, \beta}
\,,
\end{equation}
%
which we can expand out: the Minkowski metric raises the derivative with respect to \(x^{\beta }\), and we can also expand the Dalambertian and the symmetrization: we get
%
\begin{equation}
  \tensor{g}{_{\alpha \nu , \mu }^{ \alpha \mu }}
  + \tensor{g}{_{\alpha \mu , \nu }^{ \alpha \mu }}
  - \tensor{g}{_{\mu \nu, \alpha }^{ \alpha \mu }}
  =   + \tensor{g}{_{\alpha \mu , \nu }^{ \alpha \mu }}
  = \partial_{\nu }\qty( \tensor{g}{_{\mu \alpha}^{,\mu \alpha }})
\,,
\end{equation}
%
since, up to a relabeling of indices, the first and last term are equal.

The derivative \(\partial_{\nu }R\), on the other hand, is twice that: \(\partial_{\nu }R = \partial_{\nu }\qty(2 \tensor{g}{_{\mu \alpha }^{, \mu \alpha }})\), as can be directly gathered from \eqref{eq:ricci-scalar-LIF}.

Therefore, the quantity in \eqref{eq:contracted-Bianchi-LIF} is zero.

\subsubsection{Alternative expression}

Since the metric is covariantly constant, we can just bring it inside the derivative:
%
\begin{equation}
  0 = g^{\beta \mu }\nabla_{\alpha }\qty(\tensor{R}{^{\alpha }_{\mu }} - \frac{1}{2} R \delta^{\alpha }_{\mu })
  = \nabla_{\alpha }\qty(R^{\alpha \beta } - \frac{1}{2} R g^{\alpha \beta })
\,.
\end{equation}
%

These are called the \emph{Contracted Bianchi identities}, and can be alternatively derived from the Bianchi identities of the Riemann tensor, \(R_{\mu \nu [\rho \sigma ; \alpha ]} = 0\).

\subsection{Properties of the Ricci tensor}

\subsubsection{Symmetry generality}

The fact that \(T^{\mu \nu } = T^{\nu \mu }\) is frame invariant could be derived plainly from the fact that it is a tensor equation; to be more explicit we can say that under a change of coordinates with Jacobian matrix \(\Lambda \) we have \(\widetilde{T}^{\mu \nu } = \tensor{\Lambda }{^{\mu }_{\alpha }} \tensor{\Lambda }{^{\nu }_{\beta }} T^{\alpha \beta }\); when written in components like this the Jacobians commute, therefore we can just interchange them for both \(\widetilde{T}^{\mu \nu }\) and \(\widetilde{T}^{\nu \mu }\) to recover their equality. 

\subsubsection{Symmetry of the Ricci tensor.}

The expression \eqref{eq:Ricci-tensor-LIF} is manifestly symmetric: part of it is explicitly symmetrized, part is proportional to the metric, which is symmetric.

Therefore, the Ricci tensor is symmetric in any frame.

\subsection{Weak field Einstein equations}

First of all, let me recall here the general expression for the Riemann tensor: 
%
\begin{equation}
  R^{\mu }_{\nu \rho \sigma } =
  2 \Gamma^{\mu }_{[\nu| \sigma, |\rho]} + 2 \Gamma^{\alpha }_{\sigma [\nu} \Gamma^{\mu }_{\rho ] \alpha }
\,.
\end{equation}

We want to work towards Newton's equation, which is \(\nabla^2 \Phi = 4 \pi G_N \rho \), where \(\Phi \) is the gravitational field strength.

We put ourselves in the \emph{weak field} case, when \(g_{\mu \nu }= \eta_{\mu \nu } + h_{\mu \nu }\) with \(\abs{h_{\mu \nu }} \ll 1 \)

\end{document}