\documentclass[main.tex]{subfiles}
\begin{document}

\section*{7 October 2019}

% The exam for this part of the course: it might be around the end of october.

% The subscripts on last lecture \emph{were} inverted after all.

We introduce the \emph{mirror principle}: when the core contracts or expands, the envelope does the opposite.

The shell must remain at around the same temperature to maintain equilibrium: contracting the core would increase the temperature, therefore the envelope exapands.
This heuristic argument is actually derived from simulations.

The relevant time scale for oscillations is the free-fall, dynamical time scale.

We come back to the energy equation

\begin{equation}
  \pdv{L}{m} = \varepsilon - \varepsilon_\nu - \varepsilon_g
\end{equation}

we incorporate the nuclear energy generation rate and the energy lost as neutrino production into an effective energy generation rate per unit mass \(\varepsilon - \varepsilon_\nu = \varepsilon_{\text{eff}} \) and express the energy absorbed by the stellar layer as 
%
\begin{align}
 \varepsilon_g = \dv{Q}{t} = \varepsilon_{\text{eff}} - \pdv{L}{m}
\,.
\end{align}
%

This makes the meaning of this transfer equation clearer.
Using the first and second laws of thermodynamics, and recalling some thermodynamical values: the specific heat at constant volume \(c_V = \qty(\pdv{Q}{T})_V \), the equation of state exponents \(\chi_T \) and \(\chi_\rho
\) which satisfy: \(P = T^{ \chi_{T}}\) and \(P = \rho^{\chi_{\rho }}\) and the adiabatic exponents \(\Gamma_{1,2,3}\), which are defined by 
%
\begin{subequations}
\begin{align}
\Gamma_1 = \gamma_{\text{ad}} &= \qty(\pdv{\log P}{\log \rho })_{s}  \\
\frac{\Gamma_2 }{\Gamma_2 -1} = \frac{1}{\nabla _{\text{ad}}} &= \qty(\pdv{\log P}{\log T})_{s}  \\
\Gamma_3 -1 &= \qty(\pdv{\log T}{\log \rho })_{s}  \\
\,
\end{align}
\end{subequations}
%
and satisfy 
%
\begin{align}
\frac{\Gamma_1 }{\Gamma_3 -1} = \frac{\Gamma_2 }{\Gamma_2 -1}
\,.
\end{align}

These are all \emph{exponents} in some power law.
We use log values since our variables change by orders of magnitude.

We start with  the definition of the entropy differential: \(\dd{Q}\) is not an exact differential but \(\dd{Q} / T = \dd{S}\) is. So, we express it using the first law of thermodynamics: \(\dd{Q} = \dd{E} + P \dd{V}\). 
We assume the internal energy \(E\) to be a function of the volume \(V\) and of the temperature \(T\), so we will have: 
%
\begin{align}
\dd{E} = \pdv{E}{V} \dd{V} + \pdv{E}{T} \dd{T}
\,,
\end{align}
%
which we can substitute into the expression for the entropy differential: 
%
\begin{subequations}
\begin{align}
\dd{S} &= \frac{1}{T} \qty(\pdv{E}{V} \dd{V} + \pdv{E}{T} \dd{T} + P \dd{V})  \\
&= \qty(\frac{1}{T} \pdv{E}{V} + \frac{P}{T}) \dd{V} 
+ \frac{1}{T} \pdv{E}{T} \dd{T}  \\
&= \pdv{S}{V} \dd{V} + \pdv{S}{T} \dd{T}
\,,
\end{align}
\end{subequations}
%
so we have identified the partial derivatives of the entropy. By Schwarz's lemma, we then have the equality: 
%
\begin{subequations}
\begin{align}
\pdv[2]{S}{T}{V} &= \pdv[2]{S}{V}{T}  \\
\pdv{}{T} \qty(\frac{1}{T} \pdv{E}{V} + \frac{P}{T})&=
\pdv{}{V} \qty(\frac{1}{T} \pdv{E}{T})  \\
\frac{1}{T} \qty(\cancelto{}{\pdv[2]{E}{T}{V}} + \pdv{P}{T} ) 
- \frac{1}{T^2} \qty(\pdv{E}{V} + P) &=
\cancelto{}{\frac{1}{T} \pdv[2]{E}{V}{T}}  \\
 \pdv{E}{V} &= T \pdv{P}{T} - P 
\,,
\end{align}
\end{subequations}
%
but we can turn the derivatives with respect to \(V\) with ones with respect to \(\rho \propto V^{-1}\), by 
%
\begin{align}
\pdv{}{V} = \pdv{\rho }{V} \pdv{}{\rho } = - \rho^2 \pdv{}{\rho }
\,,
\end{align}
%
so after dividing through by \(\rho \) we find: 
%
\begin{align}
\rho \pdv{E}{\rho } &= -\frac{T}{\rho } \pdv{P}{T} + \frac{P}{\rho }  \\
&= - \frac{P}{\rho } \pdv{\log P }{\log T} + \frac{P}{\rho }  \\
\rho \pdv{E}{\rho }&= - \frac{P}{\rho } \qty(\pdv{\log P }{\log T} -1 )  \\
\pdv{E}{\log \rho }&= -\frac{P}{\rho } \qty(\chi_{T} - 1)
\,.
\end{align}
%

We used the fact that 
%
\begin{align}
\pdv{}{\log x} = \pdv{x}{\log x} \pdv{}{x}= x \pdv{}{x} 
\,.
\end{align}

Now, we can write the first law of thermodynamics for the specific energy density \(E(\rho, T)\): 
%
\begin{align}
\dd{Q} &= \dd{E} - \frac{P}{\rho^2} \dd{\rho }  \label{eq:heat-differential}\\
&= \pdv{E}{\log \rho } \dd{\log \rho } 
+ \pdv{E }{\log T} \dd{\log T} - \frac{P}{\rho}  \dd{\log \rho }  \\
&= \qty(- \frac{P}{\rho } \qty(\chi_{T} -1) - \frac{P}{\rho }) \dd{\log \rho } + \pdv{E}{\log T} \dd{\log T}  \\
&= - \frac{P \chi_T }{\rho } \dd{\log \rho } + \pdv{E}{\log T} \dd{\log T} 
\,,
\end{align}
%
which in the adiabatic (\(\dd{Q} = 0\)) case reduces to 
%
\begin{align}
\Gamma_3 - 1 \overset{\text{def}}{=} \pdv{\log T }{\log \rho } &= - \frac{P \chi_{T}  }{\rho } \pdv{\log T}{E} = - \frac{P \chi_T }{\rho T c_V}  \\
&=- \frac{1}{\rho } \times \underbrace{\frac{P \chi_T}{T}}_{\pdv{P}{T}} \times \underbrace{\frac{1}{c_V}}_{\pdv{T}{E}}  \\
&= - \frac{1}{\rho } \pdv{P}{E}
\,,
\end{align}
%
where we used the fact that \(c_V = \pdv*{E}{T}\). This can also be written as 
%
\begin{align}
\pdv{E}{P} = \frac{1}{\rho (\Gamma_3 -1)}
\,.
\end{align}

If we drop the hypothesis of adiabaticity, we can study the variation with respect to time of \(Q\), both when writing \(E = E(\rho , T)\) and when writing \(E = E(\rho , P)\). 
In the first case we can use equation \eqref{eq:heat-differential}, ``dividing through by \(\dd{t}\)'' (more formally, applying the differential covector equation to the vector \(\partial_{t}\)), after some manipulation we can bring out a factor \(T \pdv*{E}{T}\) to get: 
%
\begin{align} \label{eq:heat-variation-temperature-dependence}
\dv{Q}{t} = T \pdv{E}{T} \qty(\dv{\log T}{t} + \frac{\rho \pdv{E}{\rho } - \frac{P}{\rho }}{T \pdv{E}{T}} \dv{\log \rho }{t})
\,,
\end{align}
%
and similarly if we express \(E = E(\rho, P)\) we find 
%
\begin{align}\label{eq:heat-variation-pressure-dependence}
\dv{Q}{t} = P \pdv{E}{P} \qty(\dv{\log P}{t} + \frac{\rho \pdv{E}{\rho } - \frac{P}{\rho }}{P \pdv{E}{P}} \dv{\log \rho }{t})
\,.
\end{align}

We can simplify these two expressions by recalling some results from before, plus two more expression we now derive for the coefficients \(\Gamma_1\) and \(\Gamma_3 -1\): we assume adiabaticity, and get
%
\begin{align}
0=\dd{S} &= \frac{1}{T} \qty(\pdv{E}{\rho } \dd{\rho } + \pdv{E}{P } \dd{P} - \frac{P}{\rho^2} \dd{\rho })  \\
0&= \qty(\rho \pdv{E}{\rho } - \frac{P}{\rho } ) \dd{\log \rho }   + P \pdv{E}{P} \dd{\log P}  \\
\Gamma_1 = \left. \dv{\log P}{\log \rho } \right\vert_{\text{ad}} &= \frac{\frac{P}{\rho } - \rho \pdv{E}{\rho }}{P \pdv{E}{P}}
\,
\end{align}
%
and 
%
\begin{align}
0 = \dd{S} &= \frac{1}{T} \qty(\pdv{E}{\rho } \dd{\rho } + \pdv{E}{T} \dd{T} - \frac{P}{\rho^2} \dd{\rho })  \\
0&= \qty(\rho \pdv{E}{ \rho } - \frac{P}{\rho }) \dd{\log \rho } + T \pdv{E}{T} \dd{\log T}  \\
\Gamma_3 -1 = \left. \dv{\log T}{\log \rho } \right\vert_{\text{ad}} &= \frac{\frac{P}{\rho } - \rho \pdv{E}{\rho }}{T \pdv{E}{T}}
\,,
\end{align}
%


The identifications are: 
%
\begin{align}
\dv{Q}{t} = T \underbrace{\pdv{E}{T}}_{c_V} \qty(\dv{\log T}{t} + \underbrace{\frac{\rho \pdv{E}{\rho } - \frac{P}{\rho }}{T \pdv{E}{T}}}_{- (\Gamma_3 - 1) } \dv{\log \rho }{t})
\,,
\end{align}
%
and
%
%
\begin{align}
\dv{Q}{t} = P \underbrace{\pdv{E}{P}}_{1/ \rho (\Gamma_3 -1)} \qty(\dv{\log P}{t} + \underbrace{\frac{\rho \pdv{E}{\rho } - \frac{P}{\rho }}{P \pdv{E}{P}}}_{- \Gamma_1 } \dv{\log \rho }{t})
\,.
\end{align}

Finally, we substitute in the equation of energy conservation: 
%
\begin{align}
\dv{Q}{t} = \varepsilon_{\text{eff}} - \pdv{L}{m}
\,,
\end{align}
%
to find the equations for the evolution of the energy and pressure:
%
\boxalign{
\begin{subequations}
\begin{align}
  \pdv{\log P }{t} &= \Gamma_1 \pdv{\log \rho}{t} + \frac{\rho}{P} (\Gamma_3 - 1) \qty(\varepsilon_{\text{eff}} - \pdv{L}{m})  \\
  \pdv{\log T }{t} &= (\Gamma_3 - 1) \pdv{\log \rho}{t} + \frac{1}{c_V T} \qty(\varepsilon_{\text{eff}} - \pdv{L}{m})
\end{align}
\end{subequations}}

\subsection{Linear perturbation theory}

Say we have a solution for these equations, we look at linear perturbations of them.
This makes sense: the main solution is basically static on the pulsation time-scales.

The perturbed model is \(f = f(m)\),  the unperturbed one is \(f_0(m)\).
The Lagrangian perturbation is \(\delta f (m, t) = f(m, t) - f_0(m, t)\).

Let us consider specific cases for \(f\): the radial displacement is \(\delta r (m, t)\). The position of the layer at time \(t\) is \(r = r_0 + \delta r\).

We can write:
%
\begin{equation}
  r = r_0 \qty(1 + \frac{\delta r}{r_0} ) = r_0 (1+ \zeta)
  \,,
\end{equation}
%
where we define \(\zeta = \delta r / r_0 \).

In general the fractional perturbation \(\delta f / f_0\) is assumed to be \(\ll 1 \). So, \(\delta f / f_0 \sim \delta_f / f\). Formally, we only consider terms which are of first order in either perturbed function.
We will insert expressions which are functions of perturbations of all our variables, and thus get linear differential equations.

\subsubsection{Continuity equation}

Let us try the continuity equation, substituting in \(r = r_0 (1+\zeta)\) and \(\rho = \rho_0 (1 + \delta \rho / \rho_0)\).

\begin{align}
  \pdv{r}{m} &= \frac{1}{4 \pi r^2 \rho }  \\
  \pdv{}{m} \qty(r_0 (1+\zeta)) &=
  \frac{1}{4 \pi r_0^2} \qty(1+\zeta)^{-2} \qty(1 + \frac{\delta \rho}{\rho_0})^{-1}
\end{align}

and we use \((1+x)^n \approx 1 + nx\) plus the zeroth order equation: \(\pdv*{r_0 }{m} = 1 / 4 \pi r_0^2 \rho_0\). 
With these, we find:
%
\begin{equation}
  4 \pi \rho_0^2 \qty(\pdv{r_0}{m} (1+\zeta) + r_0 \pdv{\zeta}{m})
  = (1-2 \zeta) -\frac{\delta \rho}{\rho_0}
\end{equation}

We can collapse the equation into:

\begin{equation}
  \frac{\delta \rho}{\rho_0} =
  - 3 \zeta - 4 \pi r_0^3 \rho_0 \pdv{\zeta}{m}
\end{equation}

or, the density perturbation is proportional with a negative constant to the radial perturbation, plus a term proportional to \(\pdv*{\zeta}{m}\).
If there is a positive gradient of radial perturbation, the corresponding layer expands.

\subsubsection{Momentum conservation}

Let us also perturb the momentum conservation equation; the unperturbed solution will be at hydrostatic equilibrium, so \(\pdv*[2]{r_0}{t} = \pdv{r_0 }{t} =0 \), which means 
%
\begin{align} \label{eq:hydrostatic-equilibrium}
\pdv{P_0 }{m} = - \frac{Gm}{4 \pi r_0^{4}}
\,.
\end{align}

Substituting in we find that, to linear order: 
%
\begin{align}
\pdv[2]{r}{t} &= - 4 \pi r^2 \pdv{P}{m} - \frac{Gm}{r^2} \\
\pdv[2]{}{t} \qty(r_0 \qty(1 + \zeta )) 
&= 
- 4 \pi \qty(r_0 (1+z))^2 \pdv{}{m} \qty(P_0 \qty(1 + \frac{ \delta P }{P_0 })) - \frac{Gm}{r_0^2 (1 + \zeta^2)} 
\\
r_0 \pdv[2]{\zeta }{t} 
&=
- 4 \pi r_0^2 \qty(1 + 2 \zeta )
\qty(\pdv{P_0 }{m} \qty(1 + \frac{ \delta P}{P_0 }) + P_0 \pdv{}{m} \qty(\frac{ \delta P}{P_0 }))
- \frac{GM}{r_0^2} \qty(1 - 2z) 
\\
\frac{r_0}{4 \pi r_0^2} \pdv[2]{\zeta }{t} &= - \qty(1 + 2 \zeta )
\qty(\pdv{P_0 }{m} \qty(1 + \frac{ \delta P}{P_0 }) + P_0 \pdv{}{m} \qty(\frac{ \delta P}{P_0 }))
+ \qty(1 - 2z) \pdv{P_0 }{m} \marginnote{Used equation \eqref{eq:hydrostatic-equilibrium}.} \\ 
\begin{split}
\frac{1}{4 \pi r_0} \pdv[2]{\zeta }{t} &= 
- \qty(\pdv{P_0 }{m} \qty(1 + \frac{ \delta P}{P_0 }) + P_0 \pdv{}{m} \qty(\frac{ \delta P}{P_0 })) + \pdv{P_0 }{m} + \\
&\phantom{=}\, - 2\zeta \qty(\pdv{P_0 }{m} \qty(1 + \frac{ \delta P}{P_0 }) + P_0 \pdv{}{m} \qty(\frac{ \delta P}{P_0 }))
- 2 \zeta \pdv{P_0 }{m}
\end{split}
\,.
\end{align}
%


We have a purely geometric term \(-16 \pi r_0^2 \zeta \pdv*{P_0}{m} = 4 \zeta Gm /r_0^2\): when a layer moves outwards it expands.
There is a restoring force toward equilibrium.

Let us also consider the \(\log P\) form of the energy conservation equation.
We also perturb the adiabatic exponents and the \(\dv*{Q}{t} \) term.

After difficult manipulations we get back an equation which relates the changes in density and pressure to the change in energy: we manipulate until we get something which is similar to the original equation.

In general for a Lagrangian perturbation \(\delta (f^n) = n f_0^{n-1} \delta f\).
We have rules for the perturbation of product which work just like logarithmic derivatives. \(\delta\) commutes with partial derivatives.

In the radiative case with the diffusion approximation we can perturb the luminosity: \(L  \propto r^4 T^4 \kappa_R^{-1} \pdv{\log T}{m}\).

\begin{equation}
  \frac{\delta L}{L_0} = 4 \zeta + 4 \frac{\delta T}{T_0} - \frac{\delta \kappa_R}{\kappa_{R,0}} + \frac{\pdv{}{m}  \frac{\delta T}{T}}{\pdv{\log T}{m}}
\end{equation}

In the end, we have a set of four linear PDE equations (written as a 5-equation system).

These describe implicitly how the properties of the star change over time.

Pulsation usually affects mostly the outer layers of a star.

Moving on to

\section{Adiabatic oscillations}

Explioting the adiabatic approximation we will get the LAWE: Linear Adiabatic Wave Equation: a single equation which summarizes the 4.

This can be solved explicitly.

We suppose that each layer does not lose nor gain heat: \(\delta (\varepsilon_{\text{eff}} - \pdv*{L}{m} )=0\).

Is this approximation justified? The term multiplying \(\delta (\varepsilon_{\text{eff}} - \pdv*{L}{m} )\) in the perturbed energy equation is \(\rho/P (\Gamma_3-1) = \chi_T / (c_V T)\).
Usually \(\chi_T \sim 1\), \(\Gamma_1 \sim 1\).

This term,  \( \chi_T / (c_V T) \delta (\varepsilon_{\text{eff}} - \pdv*{L}{m} )\), is of the order \(1/\tau_{\text{th}}\), the thermal time scale of this layer, while the term before, \(\Gamma_1 \pdv*{}{t} (\delta \rho / \rho)\), is of the order \(1/\tau_{\text{dyn}}\), the dynamical time scale.

Therefore, we neglect the second part.
This only works for the star as a whole, not for single layers.
There are stellar layers which are \emph{strongly} non-adiabatic (driving layers). We will need some non-adiabatic theory to explain how pulsations \emph{start}.

[Insert derivation of the LAWE, which can be found in the slides].

\begin{equation}
  r \pdv[2]{\zeta}{t} =
  4 \pi r^2 \zeta \pdv{}{m} \qty((3 \Gamma_1 - 4)P)+
  \frac{1}{r} \pdv{}{m} \qty(16 \pi^2 \Gamma_1 P \rho r^6 \pdv{\zeta}{m} )
\end{equation}

We decompose: \(\zeta(m, t) = \eta(m) e^{i \sigma t}\) with a constant \(\sigma\): putting this into the LAWE we simplify the exponentials and get the space dependent form of the LAWE.
It is a Storm-Liouville equation.

\end{document}
