\documentclass[main.tex]{subfiles}
\begin{document}

\subsection{Frequency evolution and time to coalesce}

\marginpar{Friday\\ 2020-4-3, \\ compiled \\ \today}

% As we were discussing yesterday, we can approximate every orbit as a circular one. 
From the equation for the evolution of the binary's radius \eqref{eq:radius-evolution-binary}, its explicit expression in terms of \(\omega_{s}\) \eqref{eq:derivative-radius-evolution-binary} and the expression for the gravitational wave emitted power \eqref{eq:gravitational-wave-total-power-emitted} we can write 
%
\begin{align}
- \frac{2}{3} \frac{\dot{\omega}_{s}}{\omega_{s}} R &= - \frac{R^2}{GM \mu } \dot{E}_{GW} \\
\dot{\omega}_{s} &= \frac{3}{\mu } \qty(GM)^{-2/3} \omega_{s}^{1/3} 
\frac{32}{5} \frac{c^{5}}{G} 
\qty( \frac{G M_c \omega_{GW}}{2 c^{3}})^{10/3} \marginnote{Substituted in \(\dot{E}\), used \(R = (GM)^{1/3} \omega_{s}^{-2/3} \).}  \\
&= \frac{6}{5} \sqrt[3]{2} \qty(\frac{M_c G}{c^{3}})^{5/3} \omega_{GW}^{11/3}
\,.
\end{align}

If we substitute in \(\omega_{GW}\) from \(\omega_{s}\)
we finally get the relation 
%
\begin{align}
\dot{\omega}_{GW} = \frac{12}{5} \sqrt[3]{2}
\qty(\frac{M_c G}{c^{3}})^{5/3} \omega_{GW}^{11/3}
\qquad \text{or} \qquad
\dot{f}_{GW} = \underbrace{\frac{96}{5} \pi^{8/3} \qty(\frac{M_c G}{c^{3}})^{5/3}}_{k}
f_{GW}^{11/3} 
\,.
\end{align}

This can be integrated directly, by separating the variables as \(f^{-11/3} \dd{f} = \dd{t}\): the result is 
%
\begin{align}
- \frac{3}{8k} f^{-8/3} = t - t _{\text{coal}} 
\overset{\text{def}}{=} -\tau 
\,,
\end{align}
%
where \(t _{\text{coal}}\) is an integration constant, chosen so that \(t = t _{\text{coal}}\) when the frequency diverges.
We can get the frequency as a function of the time until coalescence \(\tau \):
%
\begin{align} \label{eq:gw-frequency-evolution}
f_{GW} = \qty( \frac{3}{8 k \tau })^{3/8} 
= \frac{1}{\pi } \qty( \frac{5}{256 \tau })^{3/8} \qty(\frac{c^3}{G M_c})^{5/8}
\,.
\end{align}

\todo[inline]{Wrong sign in the slides at this point.}

Inverting this expression we can get the time until coalescence from the frequency: 
%
\begin{align}
\tau = \frac{5}{256} \qty( \frac{1}{\pi f_{GW}})^{8/3} \qty( \frac{c^3}{G M_c})^{5/3}
\,.
\end{align}

Now, we want to get an expression for the \textbf{evolution of the radius}. We know that \(\omega_{GW} \propto \tau^{-3/8}\), so we can calculate 
%
\begin{align}
\frac{1}{\omega_{GW}} \dv{\omega_{GW}}{t} = 
- \frac{1}{\omega_{GW}} \dv{\omega_{GW}}{\tau } = 
+ \frac{3}{8} \frac{\tau^{-11/8}}{\tau^{-3/8} }
= \frac{3}{8} \frac{1}{\tau }
\,,
\end{align}
%
which is also equal to \(\dot{\omega}_{s} / \omega_{s}\), since it is a constant multiple of \(\omega_{GW}\). Then, by using the relation between the logarithmic derivatives of \(R\) and \(\omega_{s}\) \eqref{eq:log-derivatives-radius-angular-velocity-binary} we can write 
%
\begin{align}
\frac{\dot{\omega}_{s}}{\omega_{s}} = \frac{3}{8} \frac{1}{\tau } = - \frac{3}{2} \frac{\dot{R}}{R} 
\implies \frac{\dot{R}}{R} = - \frac{1}{4 \tau }
\,,
\end{align}
%
which is solved by 
%
\begin{align}
R(\tau ) = R_0 \qty(\frac{\tau }{\tau_0 })^{1/4} = R_0 \qty(\frac{t _{\text{coal}} - t}{t _{\text{coal}} - t_0 })^{1/4}
\,,
\end{align}
%
where \(\tau_0 \) is the time to coalescence at \(t_0 \), and \(R_0 \) is the orbital radius at \(t_0 \). 

If we plot this, it has a ``plunge'' phase near \(\tau = 0\), near which our assumptions of a nonrelativistic system break down. 

\todo[inline]{Add plot maybe}

The radius and frequency at \(R = R_0 \) are related by Kepler's third law: 
%
\begin{align}
R_0 = \sqrt[3]{ \frac{GM}{\omega_{s}^2 (\tau_0 )}}
= \sqrt[3]{ \frac{GM}{\pi^2 f_{GW}^2 (\tau_0 )}}
\,,
\end{align}
%
which can be combined with the equation for the evolution of the GW frequency \eqref{eq:gw-frequency-evolution} to yield the following expression for the time until coalescence given the orbital parameters: 
%
\begin{align}
\tau_0 = \frac{5}{256}  \frac{c^{5} R_0^{4}}{G^3M^2 \mu }
\,.
\end{align}

\subsection{Chirping waveform}

The argument of the cosine is in general denoted as \(\phi (t)\) and called the \textbf{phase}, it is defined by the relation 
%
\begin{align}
\phi (t) = \int_{t_0 }^{t} \omega_{GW}(t') \dd{t'} &= -\int_{\tau_0 }^{\tau } \dd{\tau '} \frac{2 \pi }{\pi } \qty(\frac{5}{256} \frac{1}{\tau'})^{3/8} \qty(\frac{c^{3}}{G M_c})^{5/8} \\
&= - 2 \qty( \frac{c^3}{5 G M_c})^{5/8} \tau^{5/8} + \phi_0 
\,,
\end{align}
%
which for a circular orbit is instead just a linear function of \(t\); the angular frequency is given by its constant derivative \(\omega_{GW} = \phi'\).

For a quasi-circular orbit we need to generalize, and the mass moments will look like 
%
\begin{align}
M_{11} = \frac{\mu R^2(t) \qty(1 + \cos(\phi (t)))}{2}
\qquad \text{and} \qquad
M_{12} = \frac{\mu R^2(t) \sin(\phi (t))}{2}
\,,
\end{align}
%
however since \(\dot{\omega}_{GW} \ll \omega^2_{GW}\) and \(\dot{R} \ll R  \omega_{GW}\) we will be able to neglect the derivatives of \(R\) and \(\omega_{GW}\) when computing the second time derivatives of the mass moments.
So, our two polarizations' amplitudes will read 
%
\begin{align}
h_{+} (t) &= \frac{4}{r} \qty(\frac{GM_c}{c^2})^{5/3} \qty(\frac{\pi f_{GW} (t _{\text{ret}})}{c})^{2/3} \frac{1 + \cos^2\theta}{2} 
\cos(\phi (t _{\text{ret}})) \\
h_{ \times } (t) &= \frac{4}{r} \qty(\frac{GM_c}{c^2})^{5/3} \qty(\frac{\pi f_{GW} (t _{\text{ret}})}{c})^{2/3} \cos\theta  
\sin(\phi (t _{\text{ret}})) 
\,,
\end{align}
%
which can be expressed in terms of the time to coalescence \(\tau \) as 
%
\begin{align}
h_{+} (t) &= \frac{1}{r} \qty( \frac{G M_c}{c^2})^{5/4} \qty( \frac{5}{c \tau })^{1/4} \frac{1 + \cos^2\theta }{2} \cos(\phi (\tau )) \\
h_{ \times } (t) &= \frac{1}{r} \qty( \frac{G M_c}{c^2})^{5/4} \qty( \frac{5}{c \tau })^{1/4} \cos \theta  \sin(\phi (\tau )) 
\,,
\end{align}
%
since as long as we are measuring from a fixed position the time to coalescence is the same for any observer seeing the same part of the signal: we can compute \(\tau = t _{\text{coal}} -t = \qty(t _{\text{coal}}+ \Delta t) - (t + \Delta t)\) for any fixed shift \(\Delta t\).

This model predicts a signal with a \textbf{chirping waveform}, whose amplitude increases and frequency both increase.
The chirping waveform cannot be trusted near the end, when our quasi-circular orbit approximation breaks down. 

A result from the study of the orbit of a test particle around a Schwarzschild black hole is that when the radius of the orbit goes below the Innermost Stable Circular Orbit:
%
\begin{align}
R < R _{\text{ISCO}} = \frac{6GM}{c^2}
\,
\end{align}
%
the orbit becomes unstable.
This formula holds for extreme mass ratios which can approximate the scenario of a test mass in a large masses' gravitational field. 
We have not seen any mergers like this yet, although it is definitely possible that they might exist.
 
Anyway, we can use the Schwarzschild radius as a rough guideline to see when the system becomes very relativistic and our approximations break down.

The frequency corresponding to the ISCO can be also calculated: 
%
\begin{align}
f_{GW, ISCO} = \frac{1}{6 \sqrt{6} \pi } \frac{c^3}{GM}
\,.
\end{align}

From numerical full-GR simulations and the measured GW signals we can see that the shape of the chirping waveform is basically correct even after the approximations break down; it only slightly goes out of phase with the numerical relativity calculation. 

\todo[inline]{Numerical relativity has a \emph{lower} frequency than the quadrupole approx!}

\subsection{Eccentric binaries}

We only present the results of the rather long computations; radiated power is higher than the corresponding circular binary, it is given by
%
\begin{align}
\dv{E}{t} = \frac{32}{5} \frac{G \mu^2}{c^{5}} a^{4} \omega_0^{6} f(e)
\,,
\end{align}
%
where 
%
\begin{align}
f(e) = \frac{1}{(1 - e^2)^{7/2}} \qty(1 + \frac{73}{24} e^2 + \frac{37}{96} e^{4}) \geq 1 
\,.
\end{align}

Here the parameter \(e\) is the eccentricity of the elliptical orbit: the major semiaxis is given in terms of the minor one by \(a = b / \sqrt{1 - e^2}\).
This enhancement is due to the fact that the emission is large at the point of closest approach. 

This means that the time to coalescence is \emph{shorter} for elliptical binaries: for an initial eccentricity \(e_0 \sim 1\) we have \(\tau \sim 2 \qty(1- e_0^2)^{7/2} \tau _{\text{circ}}\).

We have emission at frequencies which are integer multiples of the circular-orbit one; the peak of this spectrum shifts upward as \(e\) increases. 

GW emission has the effect of circularizing the orbit rapidly --- by the time the merger arrives it is usually quite close to circular, so we usually detect circular systems. 

\section{Hulse-Taylor binaries}


It is debatable whether the observation of this was the first observation of gravitational waves. 

\subsection{Pulsars}


This is a binary system in which one star is a pulsar. 

What is a pulsar? It's a kind of neutron star.
Not a moral judgement, but you are completely empty.

A pulsar has a large magnetic field; at a distance \(r_c = c/\omega \) the field lines cannot close so a radio beam escapes. 
This provides a clock!

``Taking the pulse'' of a pulsar: they usually have a certain well-defined shape, if we average over a few periods. 

The procedure is: take signal, FFT to get the fundamental, average over periods. 

The period can then be measured precisely, and we can observe its variations. 

Some relevant frequencies: the radio waves are on the order of \SI{e8}{Hz}, the pulsar's frequency is of the order \SI{10}{Hz}, the frequency of the binary period is \SI{e-5}{Hz}, the motion of the Earth around the Sun at \SI{e-8}{Hz} is also relevant. 

Since the pulsar frequency is very small, we can still average many pulses and still be measuring at what is basically ``a single point''. 



\end{document}
