\documentclass[main.tex]{subfiles}
\begin{document}

\marginpar{Friday\\ 2020-4-24, \\ compiled \\ \today}

% \todo[inline]{Recover ten minutes}
It is interesting to consider the fundamental mode for the oscillator, which is given by setting \(n =0\): this yields 
%
\begin{align}
\ddot{\xi}_{0} + \omega_0^2 \xi_{n} = \frac{2L}{\pi^2} \ddot{h}^{TT}_{xx}
\,,
\end{align}
%
where \(\omega_0 = \pi v_s /L\). 
The total energy in the bar can be recovered from this mode, by 
%
\begin{align}
E = \frac{M}{4} \qty(\dot{\xi}_{0}^2 + \omega_0^2 \xi_0^2 )
= \frac{1}{2} \int \dd{m} \qty(\dot{u}^2 + v_s^2 \qty(\pdv{u}{x})^2)
\,,
\end{align}
%
where \(M\) is the total mass of the bar. This basically amounts to integrating \(1/2 m v^2\), where \(v\) is calculated including both temporal and spatial variations. 

The equation for this fundamental mode has the same form as the one of a simple harmonic oscillator, whose mass is \(m_0 = M/2\), whose frequency is \(\omega_0 \) and which is subject to an effective force 
%
\begin{align} \label{eq:effective-force-harmonic-oscillator-bar-fundamental}
F(t) = \frac{2  m_0  L}{\pi^2} \ddot{h}_{xx}^{TT} = \frac{LM}{\pi^2}
\ddot{h}^{TT}_{xx}
\,.
\end{align}

Like we did in the harmonic oscillator case we introduce a damping term to model the physical imperfections in the resonance: then the equation will read 
%
\begin{align} \label{eq:damped-zeroth-mode-resonant-bar}
\ddot{\xi}_{0} + \gamma_0 \dot{\xi}_{0} + \omega_0^2 \xi_0 = \frac{2L}{\pi^2} \ddot{h}^{TT}_{xx}
\,,
\end{align}
%
where \(\gamma_0 = \omega_0 / Q_0 \). 
After Fourier-transforming we find 
%
\begin{align} \label{eq:transfer-function-resonant-bar-fundamental}
\xi_0 (\omega ) 
&= \frac{2L}{\pi^2} \frac{\ddot{h}_{xx}^{TT} (\omega )}{\omega^2- \omega_0 ^2 - i \omega \gamma_0 }  \\
&= \underbrace{-\frac{2L}{\pi^2} \frac{\omega^2}{\omega^2- \omega_0 ^2 - i \omega \gamma_0 }}_{H_{h \to \xi_0 } (\omega )}  h_{xx}^{TT} (\omega )
\,,
\end{align}
%
so we have the expression from the transfer function, allowing us to quantify the response of the system to the oscillatory stimulus of the GW. 

Typical values for the parameters at play are a length of \(L \sim \SI{3}{m}\), a mass of \(M \sim \SI{2e3}{kg}\) and a speed of sound of the order \(v_s \sim \SI{5e3}{m/s}\), which means we have a characteristic frequency of the order of \(\omega_0 \sim \SI{5e3}{rad / s}\), so the long-wavelength approximation is valid. 

The value of the decay parameter \(Q_0 \) is of the order \num{e6} typically, which means that the characteristic time for an oscillation to die out is of the order \(\tau_0 \sim 1/ \gamma_0 = Q_0 / \omega_0 \sim \num{e2} \divisionsymbol \SI{e3}{s}\): oscillations need \textbf{several minutes to die out}. 

This only applies when \(\omega \) from the GW is \textbf{very close} to \(\omega_0 \), as the transfer function is very peaked. 
% Resonant bars are resonant: their transfer function is very much peaked. 

% The transfer function is the response to the strain as a function of signal frequency.

\subsubsection{Response to periodic signals}

If we have an incoming GW propagating along the \(z\) axis, while the bar is positioned along the \(x\) axis, then the GW amplitude will be described by 
%
\begin{align}
h_{xx}^{TT} (t, z) = h_0 \Re[\exp(-i \omega (t - z/c))]
\,,
\end{align}
%
so, fixing \(z =0 \) for simplicity (and without losing generality, it is only a phase shift) its second derivative is 
%
\begin{align}
\ddot{h}_{xx}^{TT} = - \omega^2 h_0 \Re[\exp(-i \omega t)]
\,.
\end{align}

A steady-state solution to the coupled differential equation \eqref{eq:damped-zeroth-mode-resonant-bar} is:
%
\begin{align}
\xi_0 (t) &= \frac{2L \omega^2 h_0 }{\pi^2}
\Re \qty[\frac{\exp(-i \omega t)}{\omega^2 - \omega_0^2 + i \omega \gamma_0 }] \\
&= \frac{2 L \omega^2 h_0 }{\pi^2} \frac{(\omega^2 - \omega^2_{0}) \cos(\omega t) - \gamma_0 \sin(\omega t)}{(\omega^2-\omega_0^2)^2 + \omega^2 \gamma_0^2}
\,,
\end{align}
%
which has a simpler expression in Fourier space:
%
\begin{align}
\xi_0 (\omega ) = - \frac{2L}{\pi^2} \frac{\omega^2}{\omega_0^2 - \omega^2 - i \omega \gamma_0 } h_0 \delta (\omega )
\,.
\end{align}

\subsubsection{Response to bursts}

It is a fact from classical mechanics that the absorbed energy \(E\) of an oscillator with mass \(m_0 \) subject to an external impulsive force \(F(t)\) is:\footnote{The reference from Maggiore is to Landau and Liftsis, volume 1, equation 22.12.} 
%
\begin{align}
E 
= \frac{1}{2 m_0 } \abs{ \int_{ - \infty }^{ \infty  } F(t) e^{-i \omega_0 t} \dd{t}}^2
\,,
\end{align}
%
so if we suppose our force has the form \(F(t) = (2 / \pi^2) m_0 L \ddot{h}_{xx}\), where \(m_0 = M/2\) (both of these come from what we found for the equivalence to a simple harmonic oscillator of the fundamental mode of the resonant bar, \eqref{eq:effective-force-harmonic-oscillator-bar-fundamental}), we find
%
\begin{align}
E = \frac{ML^2}{\pi^{4} } \abs{ \int_{ - \infty }^{ \infty  } \ddot{h}_{xx}^{TT}(t) e^{-i \omega_0 t} \dd{t}}^2
\,,
\end{align}
%
and since the force is impulsive \(h(t)\) goes to zero quickly for \(t \to \pm \infty \), so we can integrate by parts twice: 
%
\begin{align}
E &= \frac{M L^2 (2 \pi f_0 )^{4}}{\pi^{4}} 
\abs{\int_{- \infty }^{\infty} \dd{t} h_{xx}^{TT} (t) e^{-i \omega_0 t}}^2  \\
&= 16 M L^2 f_0^{4} \abs{h_{xx}^{TT} (f_0 )}^2
\,,
\end{align}
%
since the integral is precisely the Fourier transform of \(h(t)\) evaluated at \(f_0\).
This is the energy of the GW component at frequency \(\omega_0 \): we can invert this expression to recover the energy of the component of the GW at a frequency \(f_0 \) from the energy we measure at the detector, \(E\): 
%
\boxalign{
\begin{align}
\abs{h_{xx}^{TT} (f_0 )}^2 = \frac{1}{16L^2 f_0^{4}} \frac{E}{M}
\,.
\end{align}}

\subsubsection{Ringdown}

The transfer function in \eqref{eq:transfer-function-resonant-bar-fundamental} can be written as 
%
\begin{align}
\xi_0 (\omega ) = - \frac{2L}{\pi^2} \frac{\omega^2}{(\omega - \omega_{+}) (\omega-\omega_{-})} h^{TT}_{xx} (\omega )
\,,
\end{align}
%
where the frequencies \(\omega_{\pm}\) are defined by: 
%
\begin{align}
\omega_{\pm} = \pm \sqrt{\omega_0 ^2 - \qty(\frac{\gamma_0 }{2})^2} - i \frac{\gamma_0 }{2} \approx \pm \omega_0 - i \frac{\gamma_0 }{2}
\,,
\end{align}
%
where the approximation holds when the quality factor \(Q = \omega_0 / \gamma_0 \) is large. 

\todo[inline]{The minus sign in this equation is not there in Maggiore! Should check all signs here\dots}

We suppose that the incoming burst is a delta in time: then, its spectral decomposition will be uniform.
Specifically, let us rescale it so that it has the same integral (so, the same energy) as a rectangular impulse: 
%
\begin{align}
h_{xx }^{TT} (t) = h_0 \tau_{GW} \delta (t) \implies h_{xx}^{TT} (\omega ) = h_0 \tau_{GW}
\,.
\end{align}

Then, we get 
%
\begin{align}
\xi_0 (t) &= \int \frac{ \dd{\omega }}{2 \pi } \xi_0 (\omega ) e^{-i \omega t}
\\
&= \frac{2L}{\pi^2} h_0 \tau_{GW} \int_{- \infty }^{ \infty } \frac{ \dd{\omega }}{2 \pi } \frac{\omega^2}{(\omega - \omega_{+}) (\omega-\omega_{-})} e^{-i \omega t}  \\
&\approx \frac{2L }{\pi^2} h_0 \omega_0 \tau_{GW} e^{-\gamma_0 t / 2} \sin(\omega_0 t)
\,.
\end{align}

\todo[inline]{Probably the integral is solved by looking at the complex residuals\dots }

The GW ``burst'' excites the fundamental mode, which then decays with a characteristic time \(1/ \gamma_0 \). 

By saying that the signal is a delta, do we mean that \(\tau_{GW} \ll 1 / \omega_0 \)?
Yes, where \(\omega_0 \) is the frequency of the oscillator, which is much faster than the characteristic frequency of the dissipation. 

\subsubsection{Antenna pattern}

The sensitivity of the detector depends both on the direction from which the GW comes and the polarization.

If \(\vec{J}\) is the orientation of the bar, then we have 
%
\begin{align}
h _{\text{out}} = \hat{J}^{i} \hat{J}^{j} h_{ij} (t)
\,.
\end{align}

Let us consider a ``detector-centric'' reference frame in which the bar is along the \(x\) axis, while the source is positioned in the \(xz\) plane (this is fully general, we rotate around the \(x\) axis until the condition is satisfied).
Let us also consider a ``source-centric'' reference frame in which the \(z'\) axis points towards the source of the GW, while the \(x' z'\) plane coincides with the \(xz\) one.

Since these two planes coincide, the \(\hat{y}\) vector also will: we will be able to move between the two references with a rotation around it, 
%
\begin{align}
R = \left[\begin{array}{ccc}
\cos \alpha  & 0 & \sin \alpha  \\ 
0 & 1 & 0 \\ 
- \sin \alpha  & 0 & \cos \alpha 
\end{array}\right]
\,,
\end{align}
%
where \(\alpha = \pi /2 - \theta \), and \(\theta \) is the angle between the bar and the observation direction. 

If we consider the polarizations of the GW to be expressed with respect to the primed frame: 
%
\begin{align}
h_{ij}' = \left[\begin{array}{ccc}
h_{+} & h_{ \times } & 0 \\ 
h_{ \times } & -h_{+} & 0 \\ 
0 & 0 & 0
\end{array}\right]
\,,
\end{align}
%
then by applying the rotation matrix (twice, once for each index) we will see:
%
\begin{align}
h _{\text{out}} = h_{xx} = 
R_{1k} R_{1l} h_{kl}'
= h_{+} \sin^2\theta 
\,.
\end{align}

This reference for the source is rather restrictive; we might want to align our axes so that a binary system is more easily described. 

To move to a generic source reference frame, we must a rotation of angle \(\varphi \) around the axis \(z'\) of the source system, so that the \(x' z'\) plane is free to differ from the \(xz\) one. 

With this generalization, we find:
%
\begin{align}
h _{\text{out}} (t) = h_{+}' \sin^2\theta \cos 2 \varphi + h_{ \times }' \sin^2 \theta \sin 2 \varphi 
\,.
\end{align}

\subsubsection{Resonant amplification}

The response of the bar is quite small: we cannot really measure it; however we can transfer the energy to a lighter oscillator called the \textbf{transducer}, which will move more. 
The coupled equations of motion read: 
%
\begin{align}
\ddot{\xi}_{0} + \omega_0^2 \xi_0 + \mu \omega_{t}^2 \qty(\xi_0 - \xi_t)
&= \frac{F_0}{m_0 }  \\
\ddot{\xi}_{t} + \mu \omega_{t}^2 \qty(\xi_0 - \xi_t) &= \frac{F_t}{m_t}
\,,
\end{align}
%
where the label \(t\) refers to the transducer while \(0\) is the fundamental mode of the resonant bar, and \(\mu \) is defined as the mass ratio \(m_t / m_0\). 

The force on the transducer \(F_t\) can be approximated to be zero if the experimental apparatus is well-built, since the influence of GW onto such a small mass will be negligible.

If we apply an impulsive force in the form \(F_0 = a_0 \delta (t)\) we will see a response 

We need to tune the bar so that its resonant frequency is the same as the one of the other bar. 

The ratio of masses is \(\mu \), then we will have 
%
\begin{align}
A_t = \frac{A_0 }{\sqrt{\mu }}
\,,
\end{align}
%
and we typically can choose \(\mu \lesssim \num{e-4}\), if we go lower other noise sources dominate.

\section{Thermal and readout noise}

We saw that we can represent the stages of our processing by their transfer functions \(H_{i}(\omega )\). At each stage, though, we can get noise.

This noise is then processed by the later transfer functions. 

Say we have 
%
\begin{align}
y(\omega ) = H_1 (\omega ) H_2 (\omega ) H_3 (\omega ) x(\omega )
\,,
\end{align}
%
and we have noise after each stage. 

We can have thermal noise: noisy signals are described by a deterministic part plus a stochastic noisy part.

We can model this thermal noise using a Green's function: we get 
%
\begin{align}
x(t) = \dots
\,.
\end{align}

For thermal noise, we can assume that it is completely uncorrelated: it is due to billions of atomic interactions, so it loses memory quickly.
This means that the noise is white noise: the PSD is constant in freqency. Therefore, we can show that 
%
\begin{align}
\expval{x^2(t)} \approx \frac{A}{m_0^2 \omega_0^2} \qty(1 - e^{- \gamma_0 t})
\,.
\end{align}

Then, the average total energy of the mode is 
%
\begin{align}
\expval{E(t)} 
= \frac{1}{2} m_0 \omega_0^2 \expval{x^2(t)} + \frac{1}{2} m_0 \expval{\dot{x}^2(t)}
\approx \frac{A}{2 m_0 \gamma_0 } \qty(1 - e^{-\gamma_0 t})
\,.
\end{align}

So, we get asymptotically 
%
\begin{align}
\expval{E(t)} \to 2 k_B T m_0 \gamma_0 
\,,
\end{align}
%
which is the fluctuation-dissipation theorem: the PSD of the noise is 
%
\begin{align}
S_F(\omega ) = 4 k_B T m_0 \gamma_0 
\,.
\end{align}

We can always write 
%
\begin{align}
F(\omega ) = Z(\omega ) \dot{x}(\omega )
\,,
\end{align}
%
for any linear system.
So, the FDT gives us 
%
\begin{align}
S _{\text{F, th}} (\omega ) = 4 k_B T \Re{Z(\omega )}
\,.
\end{align}

This is in terms of the PSD of the force, which can be connected to that of the velocity.

If the energy of the DoF must be asymptotically constant, there needs to be a lossy mechanism giving energy to the DoF. 

\todo[inline]{Is there a problem with many transducers?}
\todo[inline]{Could we analyze different wavelengths by modifying the frequency}

The temperatures are of the order of milliKelvin, we cool them to close to zero by isolating them and making liquid helium go through everything. 

In this case the approximation of constant temperature is quite good.

\end{document}
