\documentclass[main.tex]{subfiles}
\begin{document}

\marginpar{Monday\\ 2020-3-30, \\ compiled \\ \today}

Last time we discussed: how can we distinguish what is a GW and what is not? 

We do this by separating them by frequency. 
We cannot really measure the nature of tensor perturbation of GWs, since we are only measuring integrated effects. 

We can precisely map the effect of the GW in time, by sampling with a frequency which is much higher than the one of the GW. 

So, let us fo this formally: if we expand the Ricci tensor to second order in \(h_{\mu \nu }\) we get 
%
\begin{align}
\frac{8 \pi G}{c^{4}} \qty(T_{\mu \nu } - \frac{1}{2} g_{\mu \nu }T)
= \overline{R}_{\mu \nu } 
+ R^{(1)}_{\mu \nu } 
+ R^{(2)}_{\mu \nu } 
\,,
\end{align}
%
where the second order in \(h_{\mu \nu }\) term, \(R^{(2)}_{\mu \nu }\), has both high and low frequency components. 

This is because, when we have a term like 
%
\begin{align}
\qty(\sin(\omega_{1} t)\sin(\omega_2 t))^2
\,,
\end{align}
%
by the prostapheresis formulas we will gets terms like \(\omega_1 + \omega 2 \) and \(\omega_1 - \omega_2 \), one of which is high frequency and the other is low frequency, since we are considering a high-frequency wavepacket.

Recall, we expand in two parameters: \(h\) and \(\lambda / L_B\). Then, we have 
%
\begin{align}
\overline{R}_{\mu \nu } \sim \partial^2 \overline{g}_{\mu \nu } \sim \frac{1}{L_B^2} 
\,,
\end{align}
%
while 
%
\begin{align}
\qty[R^{(2)}_{\mu \nu }]^{\text{low frequency}} \sim \qty(\partial h)^2
 \sim \qty(\frac{h}{\lambda })^2
\,.
\end{align}

See Maggiore, page 31 for more details. 
So, the EFE low-frequency components are 
%
\begin{align}
\overline{R}_{\mu \nu } = - \qty[R^{(2)}_{\mu \nu }]^{\text{low freq}} 
+ \frac{8 \pi G}{c^{4}} \qty(T_{\mu \nu } - \frac{1}{2} g_{\mu \nu } T)^{\text{low freq}}
\,,
\end{align}
%
so we either have \(T_{\mu \nu }=0\), in which case
%
\begin{align}
\frac{1}{L_B^2} \sim \qty(\frac{h}{\lambda })^2
\,,
\end{align}
%
or 
%
\begin{align}
\frac{1}{L_B^2} \sim \qty(\frac{h}{\lambda })^2 + T_{\mu \nu } \gg \qty(\frac{h}{\lambda })^2
\,,
\end{align}
%
so we will have \(h \ll \lambda / L_B\): so we can take averages on a scale \(\ell\) such that \(\lambda_{GW} \ll \ell \ll L_B\): then we wil find 
%
\begin{align}
\overline{R}_{\mu \nu } =
- \expval{R^{(2)}_{\mu \nu }} + \frac{8 \pi G}{c^{4}} \expval{T_{\mu \nu } - \frac{1}{2} g_{\mu \nu } T} 
\,.
\end{align}

After some math (see Maggiore): we can define a stress tensor of the GW, which looks like 
%
\begin{align}
t_{\mu \nu } = - \frac{c^{4}}{8 \pi G} \expval{R^{(2)}_{\mu \nu } - \frac{1}{2} \overline{g}_{\mu \nu } R^{(2)}}
\,,
\end{align}
%
and a ``smoothed out'' SEMT of matter: 
%
\begin{align}
\expval{T_{\mu \nu } - \frac{1}{2} g_{\mu \nu } T}
\overset{\text{def}}{=}
\overline{T}_{\mu \nu } - \frac{1}{2} \overline{g}_{\mu \nu } \overline{T}
\,,
\end{align}
%
soo the equation which will hold is 
%
\begin{align}
\overline{R}_{\mu \nu } - \frac{1}{2} \overline{g}_{\mu \nu } \overline{R} = 
\frac{8 \pi G}{c^{4}} \qty(\overline{T}_{\mu \nu } + t_{\mu \nu })
\,.
\end{align}

Do note that if we work with these, the tensor which is conserved is \(\overline{T}_{\mu \nu } + t_{\mu \nu }\): 
%
\begin{align}
\nabla^{\mu } \qty(\overline{R}_{\mu \nu } - \frac{1}{2} \overline{g}_{\mu \nu }\overline{R}) = 0 = \nabla^{\mu } \qty(\overline{T}_{\mu \nu } + t_{\mu \nu })
\,.
\end{align}

How does this look like far from the source? there, we can approximate \(\overline{g} \approx \eta_{\mu \nu }\) and \(\nabla_{\mu } \approx \partial_{\mu }\). 

This \(t_{\mu \nu }\) only has 2 physical degrees of freedom: how do we gauge the others away? The Lorentz gauge plus \(h=0\) eliminates \(5\) degrees of freedom. 

When we have terms like \(h \partial \partial h\), we can integrate by parts on a sufficiently large volume to turn them into \(\partial(h \partial ) - \partial h \partial h\). 
Using the facts \(\partial^{\mu }h_{\mu \nu } = h = \square h_{\mu \nu } = 0\) we can simplify several terms: in the end we get 
%
\boxalign{
\begin{align}
t_{\mu \nu } = \frac{c^{4}}{32 \pi G} \expval{\partial_{\mu } h_{\alpha \beta } \partial_{\nu } h^{\alpha \beta }}
\,.
\end{align}}

This is actually coordinate-independent, it can be computed in any frame we like. In the TT-gauge, we will have 
%
\begin{align}
t^{00 }= \frac{c^2}{16 \pi G} \expval{\dot{h}^2_{+} + \dot{h}^2_{ \times }}
\,.
\end{align}

Also, we will have \(t^{01} = t^{02} =0 \) by symmetry, and also \(t^{03}= t^{00}\). 

Then, if we are far enough away from the source we can compute 
%
\begin{align}
\dd{E} &= \dd{A} c \dd{t} \frac{c^2}{32 \pi G} \expval{\dot{h}^{TT}_{ij}\dot{h}^{TT}_{ij}}
\frac{ \dd{E}}{ \dd{t} \dd{A}} = \frac{c^3}{16 \pi G} \expval{\dot{h}^2_{+} + \dot{h}^2_{ \times }}
\,.
\end{align}
%
so, in order to get the total power \(\dv*{E}{t}\) we can integrate this expression in \(R^2 \dd{\Omega }\). 
In order to compute the momentum carried away, we can do a similar thing, and get the momentum flux. 

In order to do this calculation, we can use the explicit expression for the \(\Delta_{ij,kl}\) projection tensor. We find 
%
\begin{subequations}
\begin{align}
\dv{E}{t} &= \frac{r^2c^{3}}{32 \pi G} \int \dd{\Omega }
\expval{
\frac{1}{r} \frac{2G}{c^{4}} \Lambda_{ij,kl} \dot{\ddot{Q}}^{kl}
\frac{1}{r} \frac{2G}{c^{4}} \Lambda_{ij,kl} \dot{\ddot{Q}}^{kl}
}  \\
&=\frac{G}{8 \pi c^{5}} \int \dd{\Omega }
\Lambda_{ij, kl} \expval{\dot{\ddot{Q}}^{ij} \dot{\ddot{Q}}^{kl}}
\,,
\end{align}
\end{subequations}
%
where the only expression depending on the angle is \(\Lambda_{ij, kl} \): then we integrate and find 
%
\begin{align}
\dv{E}{t} = \dots
\,.
\end{align}

When we do the same from the momentum density, we get an integral in the form 
%
\begin{align}
\dv{P^{k}}{t} \propto \int \dd{\Omega } \ddot{Q} \partial^{k} \dot{\ddot{Q}}
\,,
\end{align}
%
but the integrand is odd under spatial inversion, so there is no contribution! This is not true if we go beyond the quadrupole, instead full GR calculations/simulations show that there are kicks at the merger. 

We can calculate the angular distribution in a relatively simple way, since it is easy to go to TT gauge at a point. 

The energy lost by the source at \(t _{\text{ret}} = t - r/c \) is the same as the energy measured in GW. 
\todo[inline]{Clarify: why are the two expressions calculated at the same time?}

If we model the back-reaction as a force, we have 
%
\begin{align}
\dv{E _{\text{source}}}{t} = 
\expval{\int  \dd[3]{x} \dv{F_i}{V} \dot{x}_{i}} = 
= - \frac{G}{5 c^{5}} \expval{\dv{Q_{ij}}{t} \dv[5]{Q_{ij}}{t}}
\,.
\end{align}

Equating terms and making some considerations, we get 
%
\begin{align}
\dv{F_{i}}{V} = - \frac{2G}{5c^{5}} \dv[5]{Q_{ij}}{t} \rho (t, \vec{x}) x_{j}
\,,
\end{align}
%
so finally 
%
\begin{align}
F_{i} = - \frac{2G}{5c^{5}} \dv[5]{Q_{ij}}{t} m \overline{x}_{j}
\,,
\end{align}
%
where \(\overline{x}_{j} \) is the center-of-mass coordinate. 

Then, we can calculate the torque explicitly: we get 
%
\begin{align}
T_{i} = - \frac{2G}{5c^{5}} \epsilon_{ijk} Q_{il} \dv[5]{Q_{kl}}{t}
\,,
\end{align}
%
so if we take the average we get 
%
\begin{align}
\expval{\dv{L_{i}}{t}} = -\frac{2G}{5c^{5}} \epsilon_{ijk} 
\expval{\dot{\ddot{Q}}_{jl} \dot{\ddot{Q}}_{kl}}
\,.
\end{align}

\section{Back-reaction and the evolution of binary systems}

We use the reduced mass formalism, the amplitude is 
%
\begin{align}
A = \frac{4 G^{5/3} \omega_{s}^{2/3} \mu M^{2/3}}{r c^{4}}
\,,
\end{align}
%
so we have the polarizations 
%
\begin{subequations}
\begin{align}
h_{+} &= A \frac{1 - \cos^2\theta   }{2} \cos(2 \omega_{s} t _{\text{ret}} + 2 \varphi ) \\ 
h_{\times } &= A \cos \theta  \sin(2 \omega_{s} t _{\text{ret}} + 2 \varphi )  
\,,
\end{align}
\end{subequations}
%
we define the chirp mass: 
%
\begin{align}
M_c = \mu ^{3/5} M^{2/5} = \frac{\qty(m_1 m_2 )^{3/5}}{(m_1 + m_2 )^{1/5}}
\,,
\end{align}
%
also the frequency of the GW is twice the frequency of the orbit. 

Coming back to the radiated power: it is 
%
\begin{align}
\dv{E}{t} = \frac{G}{5 c^{5}} \expval{\dot{\ddot{M}}_{ij} \dot{\ddot{M}}_{ij} - \frac{1}{3} \qty(\dot{\ddot{M}}_{kk})^2}
\,,
\end{align}
%
so we can calculate this explicitly for our binary: we need to average over a period, so we get a factor \(\expval{\sin^2\varphi } = 1/2\): 
%
\begin{align}
\dv{E}{t} = \frac{32}{5} \frac{c^{5}}{G} \qty(\frac{GM_c \omega_{GW}}{2 c^{3}})^{ 10 /3}
\,,
\end{align}
%
where we used the fact \(R^3 = GM / \omega_{s}^2\).
We can do a similar thing for the angular momentum: 
%
\begin{align}
\dv{L}{t} = \dots
\,.
\end{align}

Now, let us consider masses in quasi-circular orbit: by the virial theorem, 
%
\begin{align}
\dot{R} = - \frac{2R^2}{G m_1 m_2 } \dot{E}_{GW}
\,.
\end{align}

Therefore, as the GWs carry away energy the radius shrinks. 
But we assumed the orbits to be circular! This is fine: they are almost-circular usually. 
This is fine for most of the inspiral, until the merger phase. 

\end{document}
