\documentclass[main.tex]{subfiles}
\begin{document}

\subsection{GW interferometry in the TT gauge}

\marginpar{Friday\\ 2020-5-1, \\ compiled \\ \today}

% We were discussing the basics of GW interferometry.

The TT gauge is a coordinate system in which the mirrors are free-falling (in the \(xy\) plane at least, and for the frequencies we are considering; all the delicate considerations we must make in order to distinguish the GW from the background still hold). 

So, the mirrors are still; however the light propagating through spacetime is affected by the fact that the spacetime is ``stretched'': for our photons the metric reads
%
\begin{align}
\dd{s^2} = - c^2 \dd{t^2} + \qty(1 + h_{+} (t)) \dd{x}^2 + \qty(1 - h_{+}(t))^2 \dd{y}^2 + \dd{z}^2 = 0
\,,
\end{align}
%
so in the \(x\) arm, depending on the direction of propagation we will have: 
%
\begin{align}
\dd{x} = \sqrt{ \frac{c^2 \dd{t^2}}{1 + h_{+}(t)}}
\approx \pm c \dd{t} \qty(1 - \frac{1}{2} h_{+}(t))
\,.
\end{align}

We can integrate this along the photon's path to recover the effective travel length: for the first leg of the journey we get
%
\begin{align}
L_{x} = \int_{0}^{L_x} \dd{x} = \int_{t_0 }^{t_1 } c \dd{t} = c (t_1 - t_0 ) 
- \frac{c}{2} \int_{t_0 }^{t_1 } \dd{t} h_{+}(t) 
\,,
\end{align}
%
and for the second leg: 
%
\begin{align}
- L_x = \int_{L_x}^{0} \dd{x} = - \int_{t_1 }^{t_2 } c \dd{t} \qty(1 - \frac{h_{+}(t)}{2}) = - c (t_2 - t_1 ) + \frac{c}{2} \int_{t_1 }^{t_2 } \dd{t} h_{+}(t) 
\,,
\end{align}
%
where \(t_{0, 1, 2}\) are the times at which the photon leaves the beamsplitter, bounces off the mirror, returns to the mirror. 
so the time difference is given by 
%
\begin{align}
t_2 - t_0 = \frac{2L_x}{c} + \frac{1}{2} \int_{t_0 }^{t_2} \dd{t} h_{+}(t) 
\,,
\end{align}
%
where we can consider \(t_2 \approx t_0  + 2 L_x  /c = \tau\) in the integration bound, since the correction would be second order. 

Let us then compute the integral: 
%
\begin{align}
\frac{1}{2} \int_{t_0 }^{\tau } h_+ (t) \dd{t} = \frac{1}{2} \int_{t_0 }^{\tau } \dd{t} h_0 \cos(\omega_{GW}t) = 
\frac{h_0 }{2 \omega_{GW}} \eval{\sin(\omega_{GW} t)}_{t=t_0 }^{t = \tau } 
\,.
\end{align}

We can simplify the expression making use of the trigonometric relation 
%
\begin{align}
\sin(\alpha + 2 \beta ) - \sin(\alpha ) = 2 \sin(\beta ) \cos(\alpha + \beta )
\,,
\end{align}
%
where our \(\alpha = \omega_{GW} t_0 \), and \(\beta = \omega_{GW} L_x / c\).
This yields:
%
\begin{align}
t_2 - t_0 = \frac{2L_x}{c} + \frac{L_x}{c} h_{+} \qty(t_0 + \frac{L_x}{c}) \frac{\sin(\frac{\omega_{gw} L_x}{c})}{\frac{\omega_{gw}L_x}{c}}
\,,
\end{align}
%
while for the \(y\) arm the sign of the corrective term is inverted --- the travel time diminishes. 
Note that in this last expression we are computing \(h_+\) at the time \(t_0 + L_x / c\). 

We have divided and multiplied by \(L_x / c\) in order to recover a sinc function, \(\operatorname{sinc} x = \sin x / x\).
The argument of this function is the ratio of the length of our arm to the wavelength of the GW. 
If \(L_x / c \ll T_{GW} = 2 \pi / \omega_{GW}\), the perturbation is essentially ``frozen'' during the light's travel; in the opposite limit during the travel time of the light the perturbation oscillates back and forth, cancelling out most of the effect.

We have a maximum of the effect, then, when the GW perturbation is effectively static during the time in which we are detecting: this suggests we should build \emph{short} GW interferometers. 
Keep in mind, though, that the effect size also scales with \(L_x\): a very small detector wouldn't work. 
% If the length, on the other hand, is very large the perturbation cancels out during the flight of the photon.  
We must trade off between these two. 
If we fix the frequency, making the detector longer and longer does not help: further full oscillations of the path length will not have any effect.

The light will arrive at the beamsplitter at a time \(t = t_2 \), from which we can (at least approximately) recover \(t_0 \approx t - 2L /c \).
Similarly, we can calculate the starting times of the two beams by inverting the relation and plugging in what we have just found: 
%
\begin{align}
t_0^{x} &= t - \frac{2L_x}{c} - \frac{L_x}{c} h_{+} \qty(t - \frac{L_x}{c}) \operatorname{sinc} \qty( \frac{\omega_{GW} L_x}{c}) \\
t_0^{y} &= t - \frac{2L_y}{c} + \frac{L_y}{c} h_{+} \qty(t - \frac{L_y}{c}) \operatorname{sinc} \qty( \frac{\omega_{GW} L_y}{c})
\,.
\end{align}

If the light arrives at a certain time \(t\) to the beamsplitter from both arms, then we can compute the phase difference of the beams by starting from \(t_0^{x} - t_0^{y}\):  \(\Delta \phi = \omega_{l} (t_0^{x} - t_0^{y})\). 

The interesting thing is the phase difference: if in both arms the light reaches the detector at \(t_2 \) we have 
%
\begin{subequations}
\begin{align}
\Delta \phi 
&= \omega_{l} \qty(t_0^{x }- t_0^{y}) \\
&= \omega_{l} \qty(2 \frac{L_x -L_y}{c} + \frac{2L}{c} \operatorname{sinc}(\frac{\omega_{gw} L}{c} ) h_0 \cos(\omega_{gw} t + \alpha )) = \Delta \phi_0 + \Delta \phi_{gw}
\,.
\end{align}
\end{subequations}

So, we have a tradeoff: we want to stay before the first zero of the sinc, but we also want to have a relatively large detector.

So, our optimal length is of the order of a quarter of the wavelength.

This is the same as saying we want to keep the photon in-flight for a quarter of the period of the GW. 

This would mean, for a frequency of \SI{100}{Hz}, that we would need a detector of around \SI{750}{km}.

If we do the computation accounting for the oscillation of the laser light, we get that the field out of the BS is 
%
\begin{align}
E = \frac{E_0}{2} e^{-i \gamma } \qty(e^{-i \omega_{l} t}
+ \beta e^{-i \alpha } e^{-i (\omega_{l} - \omega_{gw})} + \beta e^{i \alpha } e^{-i (\omega_{l} + \omega_{gw}) t})
\,,
\end{align}
%
which is the same of a single field which is modulated in amplitude.

\subsection{Lasers and cavities}

The best mirrors in the world are built for GW detectors.
These are dielectric mirrors.
We stack dielectric interfaces on top of each other, so that the wave coming back has constructive interference while the wave being transmitted interferes destructively with the next layer.

A \textbf{cavity} is an arrangement of mirrors such that we have a closed path for light.

Mirrors are symmetric, if we can input some light then we are also losing the same amount.

We have an incident field \(E _{\text{in}}\), a circulating field \(E_{c}\) and a transmitted field \(E_{t}\). The ``in'' mirror is labelled 1, the ``out'' mirror is labelled 2. 

So, we have 
%
\begin{align}
E_{c} = t_1 E _{\text{in}} + r_1 r_2 E_{c} e^{-ik 2L} 
\,,
\end{align}
%
so 
%
\begin{align}
E_{c} = E _{\text{in}} \frac{t_1 }{1 - r_1 r_2 e^{-ik2L}}
\,.
\end{align}

Here \(k\) is the wavevector of the electric field. The exponential is called the \emph{round-trip gain}. 

The reflected field is given by 
%
\begin{subequations}
\begin{align}
E_{r} &= -r_1 E _{\text{in}} + r_2 t_1 E_{c} e^{-ik2L}  \\
&= -E _{\text{in}} \frac{r_1 - r_2 e^{-ik 2L}}{1 - r_1 r_2 e^{-ik2L}}
\,.
\end{align}
\end{subequations}

On the other hand, the transmitted field is 
%
\begin{align}
E_{t} = t_2 E_c 
\,.
\end{align}

The circulating intensity can be calculated by the square modulus of the field, we then get 
%
\begin{align}
I_{c} = E^2 _{\text{in}} \abs{\frac{t_1 }{1 - r_1 r_2 e^{-ik2L}}}^2
\,,
\end{align}
%
so we want to minimize the denominator: this means we want \(L = n \pi / k\). 

As we increase reflectivity the peaks get narrower and higher. 

The distance in frequency between the peaks, \(c / 2L\), is called the \emph{free spectral range}. 

The \emph{finesse} is defined as the free spectral range divided by the FWHM of the peaks, 
%
\begin{align}
\mathcal{F} = \frac{\pi \sqrt{r_1 r_2 }}{1 - r_1 r_2 }
\,.
\end{align}

This is \(2 \pi  / \text{losses}\).

Using this, we can estimate the storage time of a photon inside the cavity: 
%
\begin{align}
\tau_{s} \approx \frac{L \mathcal{F}}{c \pi }
\,.
\end{align}

\end{document}
