\documentclass[main.tex]{subfiles}
\begin{document}

\chapter{The CMB}

\marginpar{Friday\\ 2020-3-13, \\ compiled \\ \today}

Today we discuss the CMB. 
This is discussed in the book Modern Cosmology \cite[chapter 3]{dodelsonModernCosmology2003}, we will follow the professor's notes, which are available in the Dropbox.

A note: in these lectures a dot will refer to conformal time derivatives only, if we differentiate with respect to cosmic time we shall write the derivative explicitly.
Let us suppose we have some particle species interacting, such as \(1 + 2 \leftrightarrow 3 +4\). 

The variation in time of the abundance of particle type \(1\), (which is given by the density times a volume: \(n_1 a^3\)) is given by the difference of the particles which are created and destroyed.
We write the formula first, and then explain it: this is given by
%
\begin{subequations}
\begin{align}
\begin{split}
a^{-3} \dv{(n_1 a^3)}{t}
&= \int \frac{ \dd[3]{p_{1}} }{(2 \pi )^3 2 E_1} \qty[\prod_{i=2}^{4}
\int \frac{ \dd[3]{p_{i}} }{(2 \pi )^3 2 E_i}] \times \\
&\phantom{=}\ 
\times (2 \pi )^{4} \delta^{(3)} (\vec{p}_{1} + \vec{p}_{2} - \vec{p}_{3} - \vec{p}_{4})
\delta (E_1 + E_2 - E_3 - E_4 )  \times \\
&\phantom{=}\ 
\times 
\abs{\mathcal{M}}^2
\qty[f_3 f_4 \qty(1 \pm f_1 ) \qty(1 \pm f_2 ) - f_1 f_2 \qty(1 \pm f_3 ) \qty(1 \pm f_4 )]
\end{split}
\,,
\end{align}
\end{subequations}
%
where:
\begin{enumerate}
  \item the delta functions account for momentum and energy conservation: energy is \emph{not conserved} in general in cosmology, \emph{but} we can use the equivalence principle to go to a reference frame which is locally Minkowski: in our description of an instantaneous process such as this, the deviations from this frame are negligible.
  \item \(\mathcal{M}\) is the \emph{invariant scattering amplitude} between the initial and final states.
  In Quantum Field Theory, in order to describe a process one assumes that the particles involved usually evolve according to their free Hamiltonians, while their interaction is described by an interaction Hamiltonian \(\hat{H} _{\text{int}}\) --- we put ourselves in the interaction picture, so that the free time evolution is factored out. The matrix element of the time evolution between the initial state \(\ket{12}\) and the final state \(\ket{34}\) is called the \(S\)-matrix element:
  %
  \begin{align}
  S_{12, 34} = \lim_{t \rightarrow \infty } \bra{12} U(t) \ket{34} 
  \,,
  \end{align}
  %
  where \(t\) is assumed to go to infinity since we are considering asymptotic initial and final states. This is in general hard to express, but it can be formally written as a Dyson series in terms of the interaction Hamiltonian. If we only consider this series to first order we get (roughly speaking):
  %
  \begin{align}
  S_{12, 34} \sim \underbrace{\braket{12}{34}}_{=0} + \bra{12} \hat{H}_{\text{int}} \ket{34}
  \,.
  \end{align}
  %
  The invariant scattering amplitude is then defined so that 
  %
  \begin{align}
  \abs{S_{12, 34}}^2 = (2 \pi)^{4} \delta^{(4)} (p^{\mu }_{1} + p^{\mu }_{2} - p^{\mu }_{3} - p^{\mu }_{4}) \abs{\mathcal{M}}^2
  \,.
  \end{align}
  %  
  Do note that \(\mathcal{M}\) is an amplitude, while \(\abs{\mathcal{M}}^2\) is a probability. 
  If we integrate across all of momentum space, as we are doing, we get the total probability of the process happening. For more details, see Peskin \cite[sections 7.2, 7.3]{peskinConceptsElementaryParticle2019}.

  Also, note that in our model the processes are always time-reversible, so \(\mathcal{M}_{12 \to 34} = \mathcal{M}_{34 \to 12}\).
  % where \(S\) is the \(S\)-matrix, while \(\hat{H}_{\text{int}}\) is the interaction Hamiltonian and \(\ket{12}\), \(\ket{34}\) are the intial and final states. 
  % The proportionality factor accounts for the fact that we want this element to be Lorentz invariant.
  \item The \(f_{i}\) are the phase space distributions of the different species: the terms including these account for the quantum statistics, we use \(-\) for fermions and \(+\) for bosons. The factors \(1 \pm f_i\) account for the quantum phenomena of Bose enhancing and Pauli blocking \cite[]{nordheimKineticMethodNew1928}. We include them here, but we will usually neglect them. 
  \item The \(2\pi \)-s account for the normalization of the deltas: if we were to discretize phase space we would not need them, but when we move from discrete sums (and Kronecker deltas) to integrals (and Dirac deltas) we need to divide by \(h^3\) for each integral over momentum; in our units however \(\hbar = 1\), therefore \(h = 2 \pi \). 
  \item The energy of each particle species is given by  \(E^2 = p^2+m^2\). 
  Why are there \(2E\) factors in the denominators? In principle, we should integrate in \(\dd[4]{p}\), however we work \emph{on shell}: a priori, the particle does whatever it wants, however solutions to the equations of motion are (very much) preferred in the path integral.
  Because of this, it is an excellent approximation to just impose this condition: we do 
  %
  \begin{align}
  \int \dd[3]{p} \int_{0}^{ \infty } \delta (E^2- p^2-m^2) 
  = \int \dd[3]{p} \int_{0}^{ \infty } \frac{ \delta (E - \sqrt{p^2+m^2})}{2E}
  \,,
  \end{align}
  %
  so we include the \(2E\) term in the denominator: the integral over the energies has already been performed, and by \(E\) we always mean \(E(\vec{p})\). 
\end{enumerate}

% \todo[inline]{Clarify definition of \(\mathcal{M}\). }

If there is no interaction, \(n_1 \propto a^{-3}\), as we expect for nonrelativistic matter. 

We are integrating over the momenta of all the particles except for \(1\) in order to account for all the possible ways the reaction could happen. 
% The \(2E_{i}\) factors in the denominator 

\todo[inline]{
Something like: 

The term for particle \(1\), \(E_1\), has a different origin: the time is related to the proper time by \(p^{0}\), which is \(E_1\).
The factor \(2\) is included for symmetry, it is indifferent if we include it or not since we can normalize the helicities \(g_{i}\). 

was mentioned during lecture, but why should we not integrate over \(\dd[4]{p_1}\) and apply the same reasoning as the other particles?
}

Typically we have kinetic equilibrium, as long as the scattering time is very short with respect to the Hubble time. 
So, we use the conventional distribution functions for Bose-Einstein or Fermi-Dirac gases,
%
\begin{align}
f_{\text{BE/FD}} = \qty(\exp(\frac{E- \mu }{T}) \pm 1)^{-1}
\,,
\end{align}
%
where the sign is a \(-\) for Bose Einstein statistics, while a \(+\) for Fermi-Dirac statistics. 
Here, \(E\) is the energy of the particle, \(\mu \) is its chemical potential, which quantifies how much the energy changes if we add particles, while \(T\) is the temperature. 

For the nonrelativistic particles (all of them, except the photons\footnote{As we will discuss in more detail later, the recombination temperature is around \SI{.3}{eV}}, much less than the mass of the lightest particles involved, electrons.) we have \(E - \mu \gg T\). 
If \(f\) becomes very small, then we can drop the terms \((1 \mp f_i)\).\footnote{What do these terms look like? We have 
%
\begin{align}
1 \mp f_i = 1 \mp \frac{1}{e^{(E-\mu ) / T} \pm 1}
= \frac{e^{(E - \mu ) / T} \pm 1 \mp 1}{e^{(E- \mu ) / T} \pm 1 }
= \frac{1}{1 \pm e^{- (E - \mu ) / T}}
\,.
\end{align}
%
} This is the Boltzmann limit. 

In theory we could not do this for photons, in practice we do it and the magnitude of the error is the same as the ratio \(\zeta (3) \approx 1.2\) to \(1\), where \(\zeta \) is the Riemann zeta function. 

Then, our distributions will be given by 
%
\begin{align}
f(E) = e^{ \mu / T} e^{- E/ T}
\,.
\end{align}

So the phase space distribution term is 
%
\begin{align}
f_1 f_2 - f_3 f_4 &= \exp(\frac{-(E_1 + E_2 ) + (\mu_1 +\mu_2 )}{T})
- \exp(\frac{-(E_3 + E_4) + (\mu_3 +\mu_4 )}{T})\\
&=\exp(- \frac{E_1 + E_2 }{T} ) \qty(e^{(\mu_3 + \mu_4 ) / T} - e^{(\mu_1 + \mu_2 ) / T})
\,,
\end{align}
%
where we used the fact that \(E_1 + E_2 = E_3 + E_4 \) by energy conservation. 

What do the chemical potentials look like?
If we were to enforce the Saha condition \(\mu_1 + \mu_2 = \mu_3 + \mu_4 \), which is equivalent to chemical equilibrium , then we would get precisely zero for our variation of species number.
This makes sense: if there is equilibrium, the number densities of the species do not change.
So, we cannot restrict ourselves to chemical equilibrium if we want to describe recombination. 

We can get relations for the chemical potentials starting from the number densities of the species which are already present:
the mean number density of species \(i\) is given by 
%
\begin{align}
n_i = \frac{g_i}{(2\pi)^3} \int \dd[3]{p} f(p)
= g_i e^{\mu_{i} / T} \int \frac{ \dd[3]{p} }{(2 \pi )^3} e^{-E_i / T}
\,,
\end{align}
%
where \(g_i\) is the number of helicity states of the particle (see the notes for the course in fundamentals of astrophysics and cosmology for more details \cite[]{tissinoFundamentalsAstrophysicsCosmology2020}). 

So, we can define 
%
\begin{align}
n_{i}^{(0)} = \eval{n_{i}}_{\mu_{i} = 0 }
\implies 
e^{\mu_i / T} = \frac{n_i}{n_i^{(0)}}
\,.
\end{align}

These quantities can be estimated if the particles are either very relativistic (\(E_i \approx p_i\)) or very non-relativistic (\(E_i \approx m_i + p^2 / 2 m_i\)): in the second case we have\footnote{We need to use a gaussian integral: in general, we have the theorem 
%
\begin{align}
\int \dd[n]{x} \exp(- \frac{1}{2} A_{ij} x_{i} x_{j}) = \sqrt{\frac{(2\pi )^{n}}{\det A}}
\,,
\end{align}
%
and in our case \(A_{ij} =  (m_i T)^{-1} \delta_{ij}\), so \(\det A = (m_i T)^{-3}\).
}
%
\begin{align}
n_{i}^{(0)}  
&= g_i \int \frac{ \dd[3]{p}}{(2 \pi)^3} e^{-E_i / T}
\approx \frac{g_i}{(2 \pi)^3} e^{-m_i / T} \int \dd[3]{p} e^{- \frac{p^2}{2m_i T}} \\
&= \frac{g_i}{(2 \pi)^3} e^{-m_i / T} \sqrt{(2 \pi)^3 (m_i T)^3}  
= g_i \qty(\frac{m_i T}{2 \pi })^{3/2} e^{-m_i / T}
\,,
\end{align}
%
while in the very relativistic case we find 
%
\begin{align}
n_{i}^{(0)}  
&= g_i \int \frac{ \dd[3]{p}}{(2 \pi)^3} e^{-E_i / T}
\approx \frac{g_i}{(2 \pi)^3} \int \dd[3]{p} e^{- p_i / T}  \\
&= \frac{g_i}{(2\pi )^3} 4 \pi T^3 \int e^{-x} x^2 \dd{x}  
= \frac{g_i}{\pi^2} T^3
\,.
\end{align}

For photons, the correct expression accounting for their quantum mechanical statistics would be this one, multiplied by the Riemann zeta function calculated at \(3\): 
%
\begin{align}
\zeta (3) = \sum _{n \in \mathbb{N} } \frac{1}{n^3} \approx 1.2
\,,
\end{align}
%
so we are wrong by about \SI{20}{\percent} in neglecting it. 

Inserting these \(n_i^{(0)}\) we get the simpler expression
%
\begin{align}
e^{(\mu_3 + \mu_4 ) / T} - e^{( \mu_1 + \mu_2 ) / T} 
=
\frac{n_3 n_4 }{n_3^{(0)} n_4^{(0)}}
- 
\frac{n_1 n_2 }{n_1^{(0)} n_2^{(0)}}
\,.
\end{align}

We can define the thermally-averaged cross section to encompass all the terms which do not depend on the number densities at a specific time: 
%
\begin{align}
\expval{\sigma v} &= \frac{1}{n_1^{(0)} n_2^{(0)}}
\prod_{i=1}^{4} \int \frac{ \dd[3]{p}}{2 E_i} 
e^{- (E_1 + E_2 ) / T} 
(2 \pi )^{4} \delta^{(4)} \qty(p^{\mu }_{1} + p^{\mu }_{2} - p^{\mu }_{3} - p^{\mu }_{4}) \abs{\mathcal{M}}^2
\,,
\end{align}
%
so the final equation is 
%
\begin{align}
a^{-3} \dv[]{}{t} (n_1 a^3)
= \expval{\sigma v} n_1^{(0)} n_2^{(0)} \qty[
\frac{n_3 n_4 }{n_3^{(0)} n_4^{(0)}}
- 
\frac{n_1 n_2 }{n_1^{(0)} n_2^{(0)}}]
\,.
\end{align}

The left hand side is typically of the order \(\sim n_1 / t \sim n_1 H\), since the order of magnitude of the age of the universe is the Hubble time \(H^{-1}\).

So, as the universe ages, the combination on the RHS must be ``squeezed to zero'': this is equivalent to the Saha equation, since it means that 
%
\begin{align}
e^{(\mu_1 + \mu_2 ) / T} = e^{(\mu_3 + \mu_4 ) / T} \implies \mu_1 + \mu_2 = \mu_3 + \mu_4 
\,.
\end{align}

This is also called ``chemical equilibrium'' by particle physicists, or nuclear statistical equilibrium by people studying Big Bang Nucleosynthesis. 

\section{Hydrogen recombination}

The process is 
%
\begin{align}
e^{-} + p \leftrightarrow H + \gamma 
\,,
\end{align}
%
so the Saha equation yields 
%
\begin{align}
\frac{n_e n_p}{n_H} = \frac{n_e^{(0)} n_p^{(0)}}{n_H^{(0)}}
\,,
\end{align}
%
and charge neutrality implies \(n_e = n_p\), \emph{not} \(n_e^{(0)} = n_p^{(0)}\). 

At this stage in evolution, there are already some Helium nuclei, but we ignore them. 

We define the ionization fraction 
%
\begin{align}
X = \frac{n_e}{n_e + H}
\,.
\end{align}

This then yields 
%
\begin{align}
\frac{1 - X_e^{n}}{X_e^{2}} = \frac{4 \sqrt{2} \zeta (3)}{\sqrt{\pi }} \eta \qty(\frac{T}{m_e})^{3/2} \exp(\epsilon_{0} / T)
\,,
\end{align}
%
where \(\epsilon_0 = m_p + m_e - m_H = \SI{13.6}{eV}\) is the ionization energy of Hydrogen. 

Then, we get that the temperature of recombination is \(T _{\text{rec}} \approx \SI{.3}{eV}\). 

The evolution of the ionization fraction is 
%
\begin{align}
\dv{X_{e}}{t} = (1-X_{e}) \beta (T) - X^2_{e} n_b \alpha^{(2)} (T)
\,,
\end{align}
%
where we defined the ionization rate 
%
\begin{align}
\beta (T) = \expval{\sigma v} \qty(\frac{m_e T}{2 \pi })^{3/2} e^{- \epsilon_0 /T}
\,,
\end{align}
%
and the recombination rate \(\alpha^{(2)} = \expval{\sigma v}\). 

The value of this can be solved numerically: the difference between this and the Saha equation is not great in the prediction in the recombination redshift; however the prediction of the residual ionized hydrogen is different: there is much more than Saha would predict. 

The universe gets reionized at \(z \gtrsim 6\); this is still under discussion. 

There are many ingredients in the interaction of the universe. We are interested in the photons: we want to predict the anisotropies in the CMB. 
There is a dipole due to the movement of the solar system through the CMB. Now, we want to see what our predictions are if we subtract this. 

[Scheme of the interactions.]

The metric interacts with everything, photons interact with electrons through Compton scattering, electrons interact with protons through Coulomb scattering, dark energy, dark matter and neutrinos  interact only with the metric.

Instead of Compton scattering, we use its nonrelativistic limit which applies here. 

Scattering between electrons and protons is suppressed since protons are very massive. The other terms in the universe affect the geometry and we could see them through this. 

There are models which include DM-DE coupling, and quintessence models, 
and models in which dark energy clusters. 

We are not going to consider these. 

We go back to first principles: 
%
\begin{align}
\hat{\mathbb{L}} [f] = \hat{\mathbb{C}} [f]
\,,
\end{align}
%
where \(f = f(x^{\alpha }, p^{\alpha })\), however actually we do not have that much freedom in the phase space distribution. 
If there are no collisions: \(\hat{\mathbb{L}} [f] =0 \), which is equivalent to 
%
\begin{align}
\frac{\mathrm{D}f}{\mathrm{D}\lambda } =0
\,,
\end{align}
%
where \(\lambda \) is the affine parameter. 

In the nonrelativistic case, 
%
\begin{align}
\hat{\mathbb{L}} = \pdv{}{t} + \dot{x} \cdot \nabla_{x} + \dot{v} \cdot \nabla_{v} = \pdv{}{t} + \frac{p}{m} \cdot \nabla_{x} + \frac{F}{m} \cdot \nabla_{v} 
\,,
\end{align}
%
while in the GR case we need to account for the geodesic equation: and we write 
%
\begin{align}
\dv{p^{\alpha }}{\lambda } = - \Gamma^{\alpha }_{\beta \gamma } p^{\beta } p^{\gamma }
\,,
\end{align}
%
where the affine parameter \(\lambda \) has the dimensions of a mass, in order to have dimensional consistency.

Then, the Liouville operator is 
%
\begin{align}
\hat{\mathbb{L}} = p^{\alpha } \pdv{}{x^{\alpha }}-  
\Gamma^{\alpha }_{\beta \gamma } p^{\beta } p^{\gamma } \pdv{}{p^{\alpha }} \overset{\text{def}}{=} \frac{\mathrm{D}}{\mathrm{D}\lambda }
\,.
\end{align}

This is a total derivative in phase space with respect to the affine parameter. 

In the FLRW background, \(f = f(\abs{p}, t)\) and 
%
\begin{align}
\hat{\mathbb{L}} = E \pdv{f}{t} - \frac{\dot{a}}{a} \abs{p}^2 \pdv{f}{E}
\,,
\end{align}
%
so if we define the number density 
%
\begin{align}
n(t) = \frac{g}{(2\pi )^3} \int \dd[3]{p} f(E, t)
\,,
\end{align}
%
so if we integrate over momenta we get 
%
\begin{align}
\int \frac{ \dd[3]{p}}{E} \hat{\mathbb{L}}[f] 
\,,
\end{align}
%
we find the equation from before: 
%
\begin{align}
\dot{n} + 3 \frac{\dot{a}}{a} n 
\,,
\end{align}
%
\todo[inline]{??? to check}

\subsection{Metric perturbations}

We want to write the most general perturbed cosmological metric solution to the Einstein Equations, starting from the FLRW metric: it will look like \(g_{\mu \nu } = g_{\mu \nu }^{\text{FLRW}} + h_{\mu \nu }\), for a small perturbation \(h_{\mu \nu }\).
Recall that the FLRW metric, in the zero-curvature case, is given by 
%
\begin{align}
\dd{s^2} =  - \dd{t^2} + a^2 \delta_{ij} \dd{x^{i}} \dd{x^{j}}
\,.
\end{align}
%

We will neglect spatial curvature because its effect on the \emph{geometry} of the universe is negligible; however we can see its effect in the dynamics.

In general this perturbation will be gauge-dependent, and we want to write only a number of independent components corresponding to the number of physical degrees of freedom. This is done via the scalar-vector-tensor decomposition \cite[section 2.1]{bertschingerCosmologicalPerturbationTheory2000a}.

We decompose the components of \(h_{\mu \nu }\) as: 
%
\begin{align}
h_{00} &= -2 \Phi \,, &
h_{0i} &= a \omega_{i} &
h_{ij} &= a^2 \qty(-2 \Psi \delta_{ij} + \chi_{ij})
\,.
\end{align}

So, we have distinguished two scalar degrees of freedom \(\Phi \) and \(\Psi \), plus a vector \(\omega_{i}\) and a traceless tensor \(\chi_{ij}\), which satisfies \(\chi^{i}_{i} = 0 \).
We can assume \(\chi_{ij}\) to be traceless since we can absorb any variations of the trace of the spatial part of the metric into \(\Psi \); this takes away one degree of freedom from the 6 of the tensor. 

This means we have  \(2+ 3 + 6 -1 = 10\) degrees of freedom, the right amount before accounting for gauge. 

Now, we make use of the Helmholtz decomposition: any vector field \(x^{i}\) can be written as \(x^{i} = x^{i}_{\perp} + x^{i}_{\parallel}\), where \(x^{i}_{\perp}\) is solenoidal (\(\nabla_{i} x^{i}_{\perp} = 0\)), while \(x^{i}_{\parallel}\) is irrotational (\(\epsilon^{ijk}\nabla_{i}x_{j} = 0\)). 

This means that we can split the three degrees of freedom of the vector field into one in \(x^{i}_{\parallel}\) --- since it is irrotational, it can be written as the gradient of a scalar function ---, and two in \(x^{i}_{\perp}\), since it is a 3D vector field for which one component is determined. Explicitly, this is 
%
\begin{align}
\omega_{i} = \underbrace{\partial_{i} \omega}_{\text{1 scalar dof}}  + \underbrace{\omega^{i}_{\parallel}}_{\text{2 vector dof}}
\,,
\end{align}
%


We have several names for a vector field \(x^{i}\) satisfying \(\nabla_{i} x^{i}\): it can be called solenoidal (since the magnetic field, in a solenoid but also anywhere else, satisfies this property), divergence-free, or transverse.
The last one can be figuratively interpreted to say that the vector is ``orthogonal to the gradient operator''; more formally, in Fourier space the condition reads \(k_{i} \widetilde{x}^{i} = 0\). 

We can do a similar thing for the tensor perturbation, but it is slightly more complicated: it it written in the form 
%
\begin{align}
\chi_{ij} = 
\underbrace{\chi_{ij}^{\parallel}}_{\text{1 scalar dof}} + 
\underbrace{\chi_{ij}^{\perp}}_{\text{2 vector dof}} + 
\underbrace{\chi_{ij}^{T}}_{\text{2 tensor dof}} 
\,,
\end{align}
%
where the derivative combination is dictated by the condition of \(\chi_{ij}\) being traceless. 
The tensor \(\chi^{\parallel}\) contains only one degree of freedom, since it can be written as 
%
\begin{align}
\chi^{\parallel}_{ij} = \qty(\nabla_{i} \nabla_{j} - \frac{1}{3} \delta_{ij}\nabla^2) \phi_{S}
\,,
\end{align}
%
while \(\chi^{\perp}_{ij}\) contains two degrees of freedom, since it can be written as the symmetrized gradient of a vector field \(S^{T}\):
%
\begin{align}
\chi^{\perp}_{ij} = 2 \nabla_{(i} S_{j)}^{T}
\,,
\end{align}
%
where the vector field is divergence-free: \(\nabla_{j} S^{j}_{T} = 0\).

The last term is fully transverse: its divergence vanishes, \(\nabla_{i} \chi^{ij}_{T} = 0\).  This term then contains two physical degrees of freedom, since it starts out from five --- it is a symmetric spatial tensor with zero trace --- but we impose 3 equations.

Let us enumerate the degrees of freedom we have, distinguishing them into ``modes'' based on how they transform:
\begin{enumerate}
  \item The \textbf{tensor mode} is given by \(\chi^{T}_{ij}\), the part of the spatial metric perturbation which cannot be obtained as a gradient. It has two degrees of freedom, and it transforms as a spin-2 field (which means it is unchanged by rotations of angle \(\pi \) about its axis). This corresponds to gravitational radiation. It is gauge-invariant.
  \item The \textbf{vector mode} is given by the transverse (divergence-free) vectors \(\omega_{i}^{\parallel}\) and \(S^{T}_{i}\). It has four degrees of freedom, two for each transverse vector. Two of these degrees of freedom are gauge, two are physical and correspond to the phenomenon of gravitomagnetism, which causes Lense-Thirring precession.\footnote{See the General Relativity notes \cite[equation 341]{tissinoGeneralRelativityNotes2020}: in order to derive the Lense-Thirring precession effect we must have a perturbation in the \(g_{03} \) component of the metric, corresponding to the vector \(\omega_{i}\) in our notation.} They transform like a spin-1 field: they must be rotated by \(2 \pi \) to come back to the starting point.
  \item The \textbf{scalar mode} is given by \(\Phi \), \(\Psi \), \(\omega\) and \(\phi_{S}\). It has four degrees of freedom, two of which are physical and two of which are gauge. They transform as a spin-0 field (so they are invariant under rotations). 
\end{enumerate}

We have separated the perturbations based on how they transform under the group of rotations. 
If we stick to linear perturbation theory, they are decoupled: they evolve independently. 

As we discussed, we can set 2 scalar fields and 2 vector fields to zero by our gauge choice: we choose the Poisson gauge, in which we set \(\phi_{S}\), \(\omega \) to zero for the scalars, and \(S_{i}^{T}\) to zero for the vectors. 

So, the full metric we get looks like:
%
\boxalign{
\begin{align}
\dd{s^2} = - e^{2 \Phi } \dd{t^2} 
+ 2 a \omega_{i} \dd{x^{i}} \dd{t} + a^2
\qty(e^{-2 \Psi } \delta_{ij} + \chi_{ij} \dd{x^{i}} \dd{x^{j}})
\,,
\end{align}}
%
where \(\omega_{i}\) is a transverse vector: \(\nabla^{i} \omega_{i} = 0\); while in \(\chi_{ij}\) we only have the tensor mode, so the tensor is traceless and transverse: we have \(\chi^{i}_{i} = 0 = \nabla^{i} \chi_{ij} \). 
% where we neglect spatial curvature (which we will do from now on). 
% This is because we would never be able to see the effect of spatial curvature in the geometry (although we could see it in the dynamics). 

% Let us describe the quantities we introduced.
% We have 10 degrees of freedom in the metric, 6 of which are physical, while 4 of which are gauge.
% We account for them like this: \(\Phi \) and \(\Psi \) are scalar, \(\omega \) is a vector, \(\chi_{ij}\) is a tensor. 

This is explained in more detail in the class by Nicola Bartolo (``early Universe''). These are GR perturbations. 

We can compare the physical spacetime and the idealized FLRW metric. 
We need to do this since we cannot solve the EFE if there is no symmetry. 
So, we say that spacetime is \emph{close} to the idealized spacetime. 

% We need a map between the physical and idealized spacetimes: this is called a \emph{gauge}. 
% Perturbations are classified with respect to their effect on FLRW. 
% In euclidean space we know scalars, vectors, tensors. 
% The perturbations will behave as such, under a change of coordinates in the Cartesian space which is the 3D space-like slice of FLRW. 

% \(\omega_{i}\) carries a 3D vector index. 
% \(\chi_{ij}\) contains the off-diagonal perturbations. 
% Let us start with  \(\omega_{i}\). In principle: Helmholtz's theorem says that we can decompose \(\omega_{i} = \omega_{i, \text{transverse}} + \partial_{i} \omega \).
% We say that the part we are interested in is the transverse one, but we still have a gradient. 

% We choose our \(\chi \) such that \(\chi^{i}_{j} = \chi^{i}_{j, i} =0\). 
% \(\chi \) could also contain vectors (objects with a vector index), as long as they are divergenceless. 

% It can also contain tensors: these are GWs. 

% So, we have 3 scalars (\(a\), \(\Psi \), \(\Phi \)), 3 components of a transverse vector (\(\omega_{i}\)), plus the divergence \(\omega \), while in the tensor we have \(6 - 3- 1 = 2\) degrees of freedom. 

% So we have \(10\) total degrees of freedom. 
% The reason we do this is that the degrees of freedom obey independent eqs.\ of motion. 

% There is gauge ambiguity in our problem: we could change the mapping between physical space and FLRW. 
% We could do a change such as \(x^{\mu } \rightarrow x^{\prime \mu } = x^{\mu } + \) a perturbation. 
% 
% This is explained better in the notes called ``GR perturbations''. 

% Gauge freedom allows us to drop 2 of the 4 scalars we have. 

% We use Poisson or longitudinal gauge, which is sometimes incorrectly called ``Newtonian gauge'' even though it is not Newtonian. 

Our next goal will be to solve the geodesic eqs.\  for the motion of the particles in this gauge. 

\end{document}
