\documentclass[main_zanardi.tex]{subfiles}

\begin{document}


% Zanardi è uno iettatore come Pauli

\paragraph{Questions}

How do we implement the POVM we described last time? That is left to the experimentalists.

\paragraph{Proof of the theorem}

This is useful because it shows us technicques, tools.

We have these \( E _{i}  \) from the POVM.

We wish to prove

\begin{equation}
  \cos ^{-1} \abs{\braket{\phi}{\psi}}
  = \cos ^{-1} \qty(\sqrt{p_i ^{\mathbb{E}} (\phi)} \sqrt{p_i(\psi)^{\mathbb{E}}})
\end{equation}

\begin{align}
  \abs{\braket{\phi}{\psi}} &= \abs{\sum_i \bra{\phi} E_i \ket{\psi}}  \\
  &\leq \sum_i \abs{\bra{\phi} \sqrt{E _{i} } \sqrt{E _{i}} \ket{\psi}}  \\
  &= \sum _{i} \abs{\braket{\hat{\phi_i}}{\hat{\psi_i}}}  \\
  &\leq \sum_i \norm{\sqrt{E_i} \ket{\phi}}\norm{\sqrt{E_i} \ket{\psi}}  \\
  &= \sum_i \sqrt{\ev{E_i}{\phi}} \sqrt{\ev{E_i}{\psi}}  \\
  &= \sum_i \sqrt{p_i ^{\mathbb{E}} (\phi)} \sqrt{p_i(\psi)^{\mathbb{E}}}
\end{align}

so we just apply the \( \cos ^{-1} \) to both sides.

\begin{equation}
  \mathbb{E} = \qty{\ketbra{\phi} | , \qty{\ketbra{\phi_i}} \text{span } \ketbra{\phi}^{\perp}}
\end{equation}

so the first term just gives us the upper bound, the other terms in the sum are 0: the inequality is saturated. We can just measure the projector associated with the state we want to know about.

\paragraph{Infinitesimal distance}

What is \( \dd _{B} \qty(\vec{p} , \vec{p} + \dd{\vec{p}}) \)?

\begin{align}
  \cos ^{-1} \qty{\sum_i \sqrt{p_i(p_i + \dd{p}_i)}}
  &= \cos ^{-1} \qty{\sum_i p_i \sqrt{1 + \frac{\dd{p}_i }{p_i}}}  \\
  &\sim \cos ^{-1} \qty(\sum_i p_i \qty(1 + \underbrace{\cancelto{0}{\frac{1}{2} \frac{\dd{p}_i}{p_i}}}_{\substack{\text{The probabilities}\\ \text{are normalized}}}     - \frac{1}{8} \qty(\frac{\dd{p_i}}{p_i})^{2} ))
\end{align}

but the term \( \sum_i \dd{p}_i \) vanishes, so we have

\begin{equation}
  \cos \dd{s} = 1 - \frac{1}{8}\frac{\dd{p}_i ^{2} }{p_i}
\end{equation}

but \( \cos \dd{s} \sim 1 - 1/2 \dd{s}^{2} _{B}  \), so

\begin{equation} \label{eq:fisher-metric}
  \dd{s}^{2}_{B}  = \frac{1}{4} \sum_i  \frac{\qty(\dd{p}_i)^{2} }{p_i}
\end{equation}

This is the \emph{Fisher metric}. More differential-geometry-like: take some \( \lambda \in \mathcal{M} = \qty{\text{manifold of control parameters of dimension N}} \), such that \( p_i = p_i(\lambda) \) and  \( \dd{p}_i = \sum_\mu (\partial_\mu p_i) \dd{\lambda_\mu} \).

Then

\begin{equation}
  g _{\mu \nu} = \frac{1}{4} \sum_i \frac{(\partial_\mu p_i)(\partial_\nu p_i)}{p_i}
\end{equation}

\begin{equation}
  \dd{s}_B ^2 = g _{\mu \nu} \dd{\lambda}_\mu \dd{\lambda}_\nu
\end{equation}

\section{Parameter estimation}

Take a random variable \( x \) distributed according to \(p_\theta (x)\) (a one-parameter distribution, with \(\theta\) as the parameter), and let \(\Theta\) be an estimator for the parameter \(\theta\): then \( \expval{\Theta} = \theta = \int  p_\theta(x) \Theta(x) \dd{x}  \).

We prove a super-famous bound. Differentiate the previous equation wrt \( \theta \).

\begin{subequations}\label{eq:classical-crao-start}
\begin{align}
  1 &= \int  p' _\theta (x) \Theta (x) \dd{x}   \\
  &= \braket{\frac{p'_\theta}{p_\theta}}{\Theta}_p \label{eq:classical-crao-start-subeq}
\end{align}
\end{subequations}

\textbf{Claim}: the notation \( \braket{f}{g}_p = \int  p_\theta f g \dd{x}  \) defines a scalar product.

\begin{greenbox}
  \textbf{Proof: }
  The axioms for a real scalar product are symmetry, linearity, and positive definiteness. The first two come from the properties of the regular product \(fg\); the interesting one to prove is the third. It is true since \(p_ \theta \geq 0\), and then \(\forall f: \int  p_ \theta f^2  \dd{x} \geq 0 \), also (modulo equality almost everywhere) \(\braket{f}{f}=0 \iff f=0\).
\end{greenbox}

We can subtract \( \theta \) from \(\Theta \rightarrow \overline{\Theta} = \Theta - \theta \) leaving the result unchanged, since it holds for any estimator: we restrict ourselves to \emph{unbiased} ones. 

% When differentiting in  \eqref{eq:classical-crao-start} the additional terms are
%
% \begin{equation}
%   -\partial _ \theta \int  p_ \theta \theta \dd{x} = - \theta \int  p' _ \theta \dd{x}
%   - \int p_ \theta \dd{x} = - 1
% \end{equation}

\begin{align}
  1 &\leq \norm{\frac{p_\theta '}{p_\theta}}^{2}_p \norm{\Theta}^2_p  \\
  &= \qty(\int  p_\theta \frac{(p_\theta ')^2}{p_\theta^2} \dd{x} )
  \qty(\int p_\theta (\Theta(x) - \theta)^2 \dd{x} )  \\
  &= F \var{\Theta}
\end{align}

Where \( F \) is just the Fischer metric. (To check: \( \bar{\Theta} = \Theta - \theta \) )

But this means \(\var{\Theta} \geq F ^{-1} \): this is the \emph{Cramer-Rao} inequality.

\subsection{Quantize it!}

Unbiased estimator \( \Tr (\rho_\theta ' \hat{\Theta}) = 1 \).

We can do \( \Tr (\rho_\theta \rho_\theta ^{-1} \rho_\theta ' \hat{\Theta}) = 1 \)

But this is noncommutative! We can do

\begin{equation} \label{eq:SLD-def}
  L_\rho (x)  = \frac{1}{2} \qty(\rho X + X \rho) = \rho'
\end{equation}

This is the Symmetric Logarithmic Derivative, SLD.
(recall the logarithmic derivative \( \rho' / \rho = \dv*{\log \rho}{x} \) ).

Now

\begin{equation}
  \frac{1}{2} \Tr \qty[\qty(\rho x + x \rho) \bar{\Theta}] = \Re \Tr (\rho X \Theta)
\end{equation}

and we can take this equation in absolute value. Then,

\begin{equation}
  1 \leq \abs{\Tr (\rho X \bar{\Theta})}^{2} = \abs{\braket{X}{\bar{\Theta}}}^{2}_{\rho}
\end{equation}

And like before

\begin{equation}
  1 \leq \norm{X}^{2} _{\rho}  \norm{\bar{\Theta}}^{2} _{\rho}
  = \Tr (\rho X^2 ) Tr(\rho (\Theta - \theta)^2)
\end{equation}

so then \( \var{\Theta} \geq 1/F_Q \).

\textbf{Claim}: take \( \rho = \sum_i \ketbra{i} \), then \( X _{ij} = 2\bra{i}\rho' \ket{j}/\qty(p_i + p_j) \).

So we can compute

\begin{align}
  F_Q &= \Tr(\rho X^2) = \sum_i p_i \ev{X^2}{i}  \\
  &= \sum _{ij} p_i \bra{i} X \ketbra{j} X \ket{i}  \\
  &= \sum _{ij} p_i \frac{2}{p_i + p_j} \bra{i} \rho' \ketbra{j} \rho' \ket{i} \frac{2}{p_i + p_j}  \\
  &= 2 \sum _{ij} \frac{\abs{\bra{i} \rho' \ket{j}}^2 }{p_i + p_j}
\end{align}

This is Quantum Fischer. It really is the result of \emph{classical} optimization.

The denominator diverges! but the numerator goes to zero quadratically (?).

\paragraph{Fischer metric for pure states}

Take \( \rho_\theta = \ketbra{\psi_\theta} \) gound-eigenstate of a many-body system.

\begin{equation}
  F_Q ^{\text{pure}} \sim \braket{\psi_\theta'}{\psi_\theta'} - \abs{\braket{\psi_\theta'}{\psi_\theta}}^2
\end{equation}

where \( \psi' = \partial_\theta \psi \). This is a way to solve the Lyapunov equation \( L_\theta (\rho) = \rho' \).

\( \Pi = \ketbra{\psi} \). Then \(\Pi ^2 = \Pi\), we differentiate it.
So \( \Pi' \Pi + \Pi \Pi' = \Pi' \): in the pure state case, the solution is then just \( \Pi ' = X \).

\textbf{Claim}: we can calculate  \( F_Q =  \Tr \qty(\rho \qty(\rho')^2)\)...

\begin{greenbox}
  \textbf{Proof}
  The derivative of \( \rho \) is \( \rho' = \ketbra{\psi}{\psi'} + \ketbra{\psi'}{\psi} \).
  We also know, by differentiating \(1 = \braket{\psi}{\psi} \), that \(\braket{\psi}{\psi'} = - \braket{\psi'}{\psi} \). Therefore \(\braket{\psi}{\psi'} = s\) is purely imaginary, \(s = -s^*\).

  Then computing \( \Tr(\rho \rho' \rho')\) (in a basis consisting of \(\ket{\psi } \) and other perpendicular vectors, whose contribution is null because of the first term \(\rho\)) we get:

\begin{subequations}
\begin{align}
   \Tr(\rho \rho' \rho')
   &= \bra{\psi} \qty(\ketbra{\psi}{\psi'} + \ketbra{\psi'}{\psi})^2 \ket{\psi}   \\
   &= \cancelto{1}{\braket{\psi}{\psi}}\braket{\psi'}{\psi}\braket{\psi'}{\psi}
   + \cancelto{1}{\braket{\psi}{\psi}}\braket{\psi'}{\psi'}\cancelto{1}{\braket{\psi}{\psi}} \nonumber \\
   &+ \braket{\psi}{\psi'}\cancelto{1}{\braket{\psi}{\psi}}\braket{\psi'}{\psi}
   + \braket{\psi}{\psi'}\braket{\psi}{\psi'}\cancelto{1}{\braket{\psi}{\psi}} \\
   &= \braket{\psi'}{\psi}^2
   + \braket{\psi'}{\psi'} + \braket{\psi}{\psi'}\braket{\psi'}{\psi}
   + \braket{\psi}{\psi'}^2  \\
   &= \braket{\psi'}{\psi'} + s^2 + (-s)^2 + s(-s)  \\
   &= \braket{\psi'}{\psi'} + s^2  \\
   &= \braket{\psi'}{\psi'} - \abs{\braket{\psi}{\psi'}}^2
\end{align}
\end{subequations}

since \(s^2 \) must be real and negative.

\end{greenbox}
\end{document}
