\documentclass[main.tex]{subfiles}
\begin{document}

% \section*{Fri Oct 25 2019}
\marginpar{Friday \\ 2019-10-25, \\ compiled \\ \today}

% Do we wish to have a part on stellar astrophysics right now, before going on with cosmology? Let him know.


% QFT must be dealt with not only at zero temperature, but also at finite temperatures.
% People started doing this in the seventies.

% The time variable is then periodic, with period \(2 \pi \beta \), where \(\beta  = 1/(k_B T)\).

% \todo[inline]{Is this connected to imaginary time?}

% This is ``in-in'' instead of ``in-out'': there are no equilibrium states before or after the interaction.

% We will simplify: we assume thermal equilibrium at any time in the evolution.
% A key point in cosmologi is the \emph{absence} of time translation invariance, therefore energy is not conserved.

% This part will come from Weinberg's book.

% We use units where \(c = k_B = \hbar = 1\). The number density of a certain species is 
%
% \begin{equation}
  % n(\dots ) = \frac{g}{(2\pi )^{3}} \int  \dd[3]{q} f(\vec{q})
% \,,
% \end{equation}
%
% where \(f(\vec{q})\) is the phase space distribution of the particles, while \(g\) is the number of helicity states of that species.
% The dots will be explained later: they are about the parametrization of the phase space distribution.

% For any species, \(E = \sqrt{m^2 + \abs{q}^2} \): this applies \emph{on shell}, for the classical equations of motion: when there are quantum fluctuations it does not hold.
% The energy density is 
%
% \begin{equation}
  % \rho (\dots) = \frac{g}{(2\pi )^3} \int  \dd[3]{q} E(\vec{q}) f(\vec{q}) 
% \,,
% \end{equation}
%
% while the pressure is 
%
% \begin{equation}
  % P (\dots) = \frac{g}{(2\pi )^3} \int  \dd[3]{q} f(\vec{q}) \frac{q^2}{3E}
% \,.
% \end{equation}
% 
% Because of isotropy, in the cases we need to consider, the dependence on \(\vec{q}\) is actually a dependence on \(\abs{\vec{q}} \equiv q \).

% What is our ansatz for the phase space distribution?  
%
% \begin{equation}
  % f(q) = \frac{1}{\exp(\frac{E-\mu }{T}) \mp 1 }
% \,,
% \end{equation}
%
% where the sign is \(-\) for bosons, \(+\) for fermions.

% The chemical potential deals with the flux of particles.
% We recover the Planck distribution when \(E = q\) and \(\mu = 0\): this tells us that photons do not have any chemical potential.
% If we measured \(\mu \neq 0 \) that would be called a \emph{spectral distortion}, but is seems like that is not the case.

% The rule for the sum of the chemical potentials only holds at equilibrium.

We can relate some chemical potentials by reactions, but not all of them: our system of equations will be degenerate, with degeneracy corresponding precisely to the globally conserved quantities (electric charge, lepton number, baryon number) which follow from the symmetry group of our theory. 
These can have any value and are conserved in any reaction,\footnote{This holds as long as the temperature is low enough: we are considering the reactions which are allowed by the Standard Model of interactions, with its symmetry group \(SU(3)_c \times SU(2)_L \times U(1)_Y\), but it is not currently known whether at higher temperatures (i.\ e.\ earlier times) this is the most general symmetry group which is spontaneously broken to the SM group. So, the statements we make only apply at relatively late times.} so they cannot be fixed by the system.

If there was a global electric charge, we'd expect global magnetic fields, but we only see them with magnitudes of the order of the \SI{1}{nT}, which gives an upper bound on the global charge of the universe.
So, any global electric charge would be quite small --- we will assume it is exactly zero. 

We can estimate the orders of magnitude for the abundances of the various particle species in the universe.
The baryon number is very small when compared to the number of photons in the universe, roughly speaking \(n_\gamma / n_b \sim \num{e10}\). 

The lepton number is harder to estimate, but it is reasonable to assume that it is quite small as well.
For slightly more detailed discussion, see the book by Weinberg \cite[before eq. 15.6.5]{weinbergGravitationCosmologyPrinciples1972}.

In the end, we can say that in the early universe \(\mu / T \ll 1\), so we can assume \(\mu \approx 0\).
This is just a reasonable simplification, which we make in order to get analytic results. 

Under this assumption the quantities characterizing the matter distribution in the universe only depend on the temperature: so, we will just write \(n(T)\), \(\rho (T)\) and \(P(T)\).

% \todo[inline]{There is a whole section before this in \cite[]{paccianiAppuntiCorsoPhysical2018}\dots}

In general, when dealing with thermodynamic problems in an expanding spacetime there is a complication: in Minkowski spacetime we have symmetry under time translations, thus it makes sense to talk about stationarity.
In an expanding universe, instead, we have no Killing vector with respect to time.
There is a competition between two evolutions, the thermodynamic evolution of the system and the expansion of the universe: we cannot truly have equilibrium!

The way to deal with this problem is: we assume that the first evolution is much faster than the other, that is, we reach thermal equilibrium on timescales that are short if compared to the expansion.
This way, we can neglect the expansion of the universe while our system reaches equilibrium.

So our problem is oversimplified: we assume thermodynamic equilibrium, which makes sense in certain periods of the life of the universe, and that allows us to embed a thermal situation into a universe which evolves in time.

% A proper treatment of this issue will include a term corresponding to the expansion in the equations of motion. 

\subsection{Entropy}

From the second principle of thermodynamics we know that the entropy in a certain volume \(V\) at temperature \(T\), denoted \(S(V, T)\) is given by: 
%
\begin{equation}
  \dd{S}  = \frac{1}{T} \qty(\underbrace{\dd{\qty(\rho(T) V)}}_{ \dd{E}}  + P(T) \dd{V})
  = \frac{1}{T} \qty(V \dd{\rho (T)} + (P(T) + \rho(T) ) \dd{V})
\,,
\end{equation}
%
since in order to get the total energy we must multiply the constant energy by the volume: \(E = \rho (T) V\). 
 
Then we can read off the partial derivatives of the entropy:
%
\begin{equation}
  \pdv{S}{V} = \frac{1}{T} \qty(\rho (T) + P(T))
  \qquad \text{and} \qquad
  \pdv{S}{T} = \frac{V}{T} \dv{\rho (T)}{T} 
\,.
\end{equation}

In order for the differential to be exact it needs to be closed, which means that the second partial derivatives need to commute (these are known as the \emph{Pfaff relations}):\footnote{They only hold in a simply connected space.}
%
\begin{align}
  \pdv[2]{S}{T}{V} &= \pdv[2]{S}{V}{T} \\ 
  \pdv{}{T}  \qty(\frac{1}{T}\qty(\rho (T) + P(T))) 
  &= \pdv{}{V} \qty(\frac{V}{T} \dv{\rho (T)}{T} )   \\
  - \frac{1}{T^2} \qty(\rho + P)
  + \frac{1}{T} \qty(\dv{\rho }{T} + \dv{P}{T})
  &= \frac{1}{T} \dv{\rho }{T}   \\
  \dv{P}{T} &= \frac{1}{T} \qty(\rho +P) 
  \marginnote{Simplified \(T^{-1}(\dv*{\rho }{T})\), multiplied by \(T\) and brought \(T^{-1} (\rho +P)\) to the other side}[-1cm]
  \label{eq:rewritten-pfaff-relations-entropy}
\,.
\end{align}

Cosmology has not entered into the picture yet, but it can by the third Friedmann equation, which can be rewritten as 
%
\begin{align}
\dot{\rho} &= - 3 \frac{\dot{a}}{a} (\rho + P)  \\
0 &= 3 \dot{a} a^2 (\rho + P) + a^3 \dot{\rho} 
\marginnote{Multiplied by \(- a^3\).}  \\
a^3 \dot{P} &= 3 \dot{a} a^2 (\rho + P) + a^3 \dot{\rho} + a^3 \dot{P} 
\marginnote{Added \(a^3 \dot{P}\) on both sides.}  \\
a^3 \dot{P} &= \dv{(a^3)}{t} (\rho + P) + a^3 \dv{(\rho + P)}{t}  \\
a^3 \dot{P} &= \dv{}{t} \qty(a^3 (\rho + P))
\,,
\end{align}
%
% %
% \begin{equation}
%   a^3 \dv{P}{t} =\dv{}{t} \qty(a^3 (P + \rho ))
% \,,
% \end{equation}
%
and these two, when put together, are equivalent to 
%
\begin{equation}
  \dv{}{t} \qty(\frac{a^3}{T} (\rho (T) + P (T) ))=0
\,,
\end{equation}
%
therefore this quantity is a constant of motion. 
Let us verify this statement: expanding the derivative we get 
%
\begin{align}
\dv{}{t} \qty(\frac{a^3}{T} (\rho + P )) 
&= \frac{1}{T} \dv{}{t} \qty(a^3 (\rho + P)) + a^3 (\rho + P) \dv{}{t} \qty( \frac{1}{T})  \\
&= \frac{1}{T} a^3 \dot{P} - a^3 (\rho + P) \frac{\dot{T}}{T^2}  \\
&= \frac{a^3}{T} \dv{P}{T} \dot{T} - a^3 (\rho + P) \frac{\dot{T}}{T}  \\
&= \frac{a^3}{T} \frac{(\rho + P)}{T} \dot{T} - a^3 (\rho + P) \frac{\dot{T}}{T} = 0
\marginnote{Used equation \eqref{eq:rewritten-pfaff-relations-entropy}.}
\,.
\end{align}

% GOT HERE

For the RW line element, the square of the determinant \(\sqrt{-g} =  a^3 \), so we get 
%
\begin{equation}
  \dv{}{t} \qty(\sqrt{-g }  \frac{\rho + P}{T}) = 0
\,.
\end{equation}

So the quantity which is differentiated is constant. If we plug this back into the differential expression for the entropy, we get: 
%
\begin{equation}
  \dd{S} = \dd \qty(\frac{(\rho + P) V}{T})
\,,
\end{equation}
%
therefore the differentiated quantities are equal up to a constant, but since \(V \propto a^3\) we get that the \emph{entropy is constant}.

\begin{equation}
  S \equiv S(a^3, T) 
  = \frac{a^3}{T} \qty(\rho + P)
  = \const
\,.
\end{equation}

Since \(\rho \propto T^{4}\) and \(\rho \propto a^{-4}\), we expect Tolman's law: \(Ta \sim 1\).

Also, for photons \(P  = \rho / 3\): so we get \(S \propto \frac{4}{3} (a^3/T) T^{4} \propto T^3 a^3 = \const\).

We only consider photons since they have a much larger number density.

\todo[inline]{Is this because the suppression is exponential while the ratio of energies is somehow polynomial?}

Now we wish to do the integrals for the thermodynamical quantities: 
%
\begin{equation}
  \int  \dd[3]{\vec{q}} = 4 \pi \int_0^{\infty}  \dd{q} q^2 
\,,
\end{equation}
%
so we get 
%
\begin{subequations}
\begin{align}
  n(T) &= \frac{g}{(2\pi^2} \int  \dd{q} q^2 f(q) \\ 
  \rho (T) &= \frac{g}{2\pi^2} \int  \dd{q} q^2 f(q) E(q) \\ 
  P (T) &= \frac{g}{6\pi^2} \int  \dd{q} q^2 f(q) \frac{q^2}{E(q)} 
\,,
\end{align}
\end{subequations}
%
which can be solved analytically in the ultrarelativistic and nonrelativistic approximations.

Ultrarelativistic means \(q \gg m\): actually we compare these to the temperature: we define \(x = q/T\), and then the energy is \(\sqrt{x^2+ m^2/T^2} \): we suppose \( m/T \ll 1\).

The momentum will not always be large, but the cases in which it is large give a much greater contribution if the temperature is large.

\paragraph{Ultrarelativistic}

We approximate \(E(q) \sim q\), so  
%
\begin{equation}
  n(T) = \frac{g}{2 \pi^2} \int _{\mathbb{R}^{+}} \dd{q} q^2 \qty(\exp(q/T) \mp 1 )^{-1} 
\,,
\end{equation}
%
%
\begin{equation}
    \rho (T) = \frac{g}{2 \pi^2} \int _{\mathbb{R}^{+}} \dd{q} q^3 \qty(\exp(q/T) \mp 1 )^{-1} 
  \,,
\end{equation}
%
\begin{equation}
    P (T) = \frac{g}{6 \pi^2} \int _{\mathbb{R}^{+}} \dd{q} q^3 \qty(\exp(q/T) \mp 1 )^{-1} 
\,,
\end{equation}
%
so we can see that in this approximation, which is equivalent to \(m \approx 0\), we get matter behaving like radiation: \(P = \rho /3\).  
The result of the integrals depends on the type of the particles, and it is 
%
\begin{equation}
  n(T) = \begin{cases}
      \frac{\xi (3)}{\pi^2} g T^3 & BE  \\
      \frac{3}{4}\frac{\xi (3)}{\pi^2} g T^3 & FD  
  \end{cases}
\,,
\end{equation}
%
and we get the proportionality to \(T^3\).
For the energy density:
\begin{equation}
    \rho (T) = \begin{cases}
        \frac{\pi^2}{30} g T^4 & BE  \\
        \frac{7}{8}\frac{\pi^2}{30} g T^4 & FD  
    \end{cases}
  \,,
\end{equation}
and for the pressure we just divide by 3.
For photons, \(g=2\) and they are bosons, so we get 
%
\begin{equation}
  \rho _\gamma  = \frac{\pi^2}{15} T^{4}
\,.
\end{equation}

\paragraph{Nonrelativistic}

Now \(m \gg T\), so we get \(E \approx m + \frac{q^2}{2m} + O((q/\sqrt{m})^{4})\) by expanding.

So, can we just substitute this? It would seem like this makes the exponential very large, so the difference between bosons and fermions becomes negligible. So, to zeroth order in (\(q^2/m\)) we get 
%
\begin{equation}
  f \approx \exp(-\frac{m- \mu }{T}) 
\,,
\end{equation}
%
and then 
%
\begin{equation}
  n = \frac{g}{2 \pi^2} \exp(-\frac{m- \mu }{T}) \int _{\mathbb{R}^{+}} \dd{q}q^2 
\,,
\end{equation}
%
which diverges. This is the ultraviolet catastrophe: we need the \emph{Planckian}, by considering the first order in \(q^2/m\): then
%
\begin{equation}
    n = \frac{g}{2 \pi^2} \exp(-\frac{m- \mu }{T}) \int _{\mathbb{R}^{+}} \dd{q}q^2 \exp(-\frac{q^2}{2mT}) 
    = g \qty(\frac{mT}{2 \pi })^{3/2} \exp(\frac{\mu - m}{T}) 
  \,,
\end{equation}
where we applied the identity 
%
\begin{equation}
  \int _{\mathbb{R}} \dd{x} x^2 \exp(-\alpha x^2) = \frac{\sqrt{\pi } }{2 \alpha^{3/2}} 
\,.
\end{equation}
%

The exponential factor is known as the Boltzmann suppression factor.

To recover the energy density and pressure we could do \(\rho = m n\) and \(P = n T\). 

\todo[inline]{Do these come out of doing the other two integrals?}

Therefore, \(P = T \rho / m\), so \(P \ll \rho \): for nonrelativistic particles, we can deal with them as \emph{noninteracting dust}.

\(P = nT\) is just the ideal gas law.

If we compare relativistic particles to nonrelativistic ones, the former dominate the latter in terms of all of these three quantities.

The physical consequence is: whenever temperature drops down to the mass of a certain particle, that particle becomes non relativistic and is forced to obey the Boltzmann distribution, which contains the Boltzmann suppression factor. The only way to obey this is to annihilate with its own antiparticle, thus producing radiation.



\end{document}