\documentclass[main.tex]{subfiles}
\begin{document}

\section*{8 October 2019}

We visit \url{https://heasarc.gsfc.nasa.gov}

This is the archive where we will spend most of our time: one can find the data from all the current missions.

At the link \url{http://adsabs.harvard.edu/abstract_service.html} one can find Astrophysics papers, with full text sources.

There is the extragalactic database \url{https://ned.ipac.caltech.edu/}, and a galactic database at \url{http://simbad.u-strasbg.fr/simbad/}.

In \url{https://heasarc.gsfc.nasa.gov} one can go to Tools \(\rightarrow\) Coordinate converter to find coordinates.
If you look at an object and then at another it tells you what the angular distance between them is.

In Tools \(\rightarrow\) energy converter you can convert energies.

Now, let us talk about \emph{specific intensity}:
it is defined as
%
\begin{equation}
  I_\nu = \frac{\dd{E_\nu}}{\dd{A} \dd{\Omega} \dd{\nu} \dd{t} \cos \theta}
  = \frac{h\nu\dd{N_\nu}}{\dd{A} \dd{\Omega} \dd{\nu} \dd{t} \cos \theta}
  = nh\nu\,,
\end{equation}
%
where \(n\) is the \emph{photon intensity}.

We also define \emph{flux density}, measured in \SI{}{\erg\per\square\centi\meter\per\hertz\per\second}:
%
\begin{equation}
  F_\nu = \int_{\text{source}} I_\nu \cos \theta \dd{\Omega} = \dv{N}{E}
\end{equation}
%
and \emph{flux}, measured in \SI{}{\erg\per\square\centi\meter\per\second}:
%
\begin{equation}
  F = \int_{\nu_1}^{\nu_2} F_\nu \dd{\nu}\,.
\end{equation}

All detectors are just energy-dependents photon counters.

Sometimes, the flux density follows a power law: \(F_\nu = k E^{-\Gamma}\).

\subsection{The Earth's atmosphere}

From Earth we can observe visible, near infrared \& radio light.

Far infrared, long radio waves, and from ultraviolet onwards the light is blocked.

If \(z\) is the angle between the zenith and the observation direction, we say that we have \(1 / \cos z \) airmasses to see through.
The intensity is:
%
\begin{equation}
  I(z) = I_0 e^{-\tau} = I_0 e^{\tau_0 \sec z}
\end{equation}

We can distinguish stars from planets: stars twinkle because of the seeing effect, while planets are extended for us so they do not.

\end{document}
