\documentclass[main.tex]{subfiles}
\begin{document}

\subsection{Direct measurements}

\marginpar{Thursday\\ 2022-1-20}

BESS is a balloon, typically balloons fly around \SI{40}{km}. 
The density decreases exponentially, but there we still have atmosphere
there on the order of \SI{0.5}{\percent} in terms of column density. 

Aerogel is used to measure Cherenkov light. 

Balloons typically fly for about a week. 
Balloon missions are usually focused on very low energies.

The maximum detectable rigidity for AMS is about \SI{1}{TV}. 

The flux of protons and Helium is higher than expected at high energies, 
more than a few hundreds of GV. 
The break from theory seems to be around the same rigidity for different nuclides. 

Calorimeters are required to go to higher energies. 

DAMPE saw a hardening in the spectrum around \SI{500}{GeV}, and then a softening 
around \SI{14}{TeV}. 

HERD will be a detector on the Chinese space station.

Up to the ankle it is hard to do direct measurements. 

Most muons in a shower come from pion and kaon decay. 
If the parent particle's energy increases, its decay length will increase, 
therefore the ratio of interaction length to the decay length will decrease,
therefore the probability of interaction as opposed to decay increases, 
therefore the muon flux decreases. 

If we are able to distinguish the number of electrons and the number of muons 
in the shower, we can distinguish showers in which the primary was more massive:
an iron-sourced shower has more muons per electron. 

The most difficult region is across the knee:: \SI{e12}{eV} to \SI{e17}{eV},
since we need to stay on the ground, but we do not have fluorescence light available.

Electrons and positrons from the shower spiral around the geomagnetic field
if they are low-energy enough. 

The observable is \(\log A\). 


\end{document}