\documentclass[main.tex]{subfiles}
\begin{document}

\marginpar{Tuesday\\ 2021-11-16}

Last time we compute the threshold for reactions with a CMB photon interacting with some high-energy particles. 

We can plot the mean free path with varying energy. 
It is roughly infinite below a certain threshold (we can have some interactions with the tails of the distribution near it), then it asymptotes to \(\Lambda \sim \SI{100}{Mpc}\). 

Around \SI{5e19}{eV} the \(\ce{p} + \gamma _{\text{CMB}} \to \Delta^{\pm}\) reaction kicks in, while at around \SI{e14}{eV} we get  the \(\gamma + \gamma  _{\text{CMB}}\to e^{+} e^{-}\) reaction. 
There, however, the energy affects the cross-section of the reaction, so we expect to see a minimum. 

The all-particle energy spectrum has an ultra-high-energy cutoff around \SI{e20}{eV} because of these reactions. 

At LHC we have center-of-mass energies of \SI{14}{TeV}, which is found by adding together the energies of the bins.
This is true in this particular case since we have a symmetric configuration, meaning that the total momentum in the lab frame as well vanishes. 

In a fixed target context (such as a cosmic ray coming in) instead we get 
%
\begin{align}
\sqrt{s} = \sqrt{2 m (E + m)}
\,,
\end{align}
%
so the correct comparison reads 
%
\begin{align}
\sqrt{2 m (E +m)} = 2 E^{*}
\,,
\end{align}
%
where \(E^{*} = \SI{7}{TeV}\). Therefore, 
%
\begin{align}
E = \frac{(2 E^{*})^2}{2m} - m
\,.
\end{align}

It depends on \(E^{*}\) quadratically! 
In our case, with protons, we get \(E \sim \SI{e17}{eV}\).

How are the angles changed? 
The invariant mass of the total system is 
%
\begin{align}
m _{\text{inv}} = \sqrt{\qty(\sum _{i}E_i)^2 - \abs{ \sum _i p_i}^2}
\,.
\end{align}

We also know that \(\gamma = E / m\).
The total \(\gamma \) can be computed by looking at this expression with the total energy \(E _{\text{tot}}\) and \(\sqrt{s} = m _{\text{inv}}\) as the total mass. 

In the case of our cosmic rays, this reads 
%
\begin{align}
\gamma _{\text{CMS}} \sim \sqrt{ \frac{\SI{e17}{eV}}{\SI{2e9}{eV}}} \approx \SI{7e3}{}
\,,
\end{align}
%
which corresponds to an angle of \(1 / \gamma _{\text{CMS}}\sim \SI{30}{\arcsec}\). 

This is a good thing, in a way: instead of having to make a detector all around the event, we can make a smaller one in the forward direction only. 

We can define a new kinematic variable in order to better understand these angles. 

Consider a Lorentz transformation with a velocity \(v_0 = \beta_0 c\) along the \(x\) axis. 
The transformation for a velocity \(v\) reads 
%
\begin{align}
v = v' + v_0 
\,
\end{align}
%
in the NR case, while for the relativistic case it will be 
%
\begin{align}
v = \frac{v' + v}{1 + v' v_0 / c^2}
\,.
\end{align}

This only holds for the \(x\) component, along the boost.

Therefore, the velocities are not additive.
A trick is to introduce a new quantity, the \emph{rapidity}:
%
\begin{align}
y' = \frac{1}{2} \log \frac{1 + \beta '}{1 - \beta '} = \operatorname{arctanh} (\beta ')
\,,
\end{align}
%
and rapidities are indeed additive. 
These are typically considered only looking at the component parallel to the boost: 
%
\begin{align}
y = \operatorname{arctanh} (\beta_{\parallel})
\qquad \text{where} \qquad
\beta_{\parallel} = \beta \cos \theta = \frac{p_{\parallel}}{E}
\,.
\end{align}

This formulation holds in 3D as well. 
The rapidity can also be written as 
%
\begin{align}
y = \frac{1}{2} \log \qty(\frac{E + p_{\parallel}}{E - p _{\parallel}})
\,.
\end{align}

The Lorentz parameters can be written in terms of hyperbolic functions: 
%
\begin{align}
\beta &= \tanh(y)  \\
\gamma &= \cosh(y)  \\
\beta \gamma &= \sinh(y)
\,,
\end{align}
%
so that the Lorentz matrix reads 
%
\begin{align}
\left[\begin{array}{cc}
\gamma  & - \beta \gamma  \\ 
- \beta \gamma   &  \gamma 
\end{array}\right]
= \left[\begin{array}{cc}
\cosh(y) & -\sinh(y) \\ 
-\sinh(y) & \cosh(y)
\end{array}\right]
\,.
\end{align}

In the ultrarelativistic limit, the rapidity reads 
%
\begin{align}
y &\approx \log \sqrt{\frac{1 + \cos \theta }{1 - \cos \theta }} \\
&\approx \log \frac{\sqrt{(1 + \cos \theta ) / 2}}{\sqrt{( 1 - \cos \theta ) / 2}}  \\
&\approx - \log \tan(\theta / 2 ) = \eta 
\,,
\end{align}
%
where \(\eta \) is called the pseudo-rapidity. 

\todo[inline]{make figure}

At LHC we can only detect particles which are not emitted very forward or very backward. 

What about the rapidities of the particles for the case of a \SI{e17}{eV} cosmic ray? 
They are additive, so \(y _{\text{lab}} = y' + y_0 \), where 
%
\begin{align}
y_0 &= \frac{1}{2} \log \qty(\frac{1 + \beta _{\text{CMS}}}{1 - \beta _{\text{CMS}}})  \\
\beta _{\text{CMS}} &= \frac{\abs{p'}}{E+m} \approx \frac{E}{E+m} \\
y_0 &\approx \frac{1}{2} \log \qty(\frac{2E + m}{m}) \approx \frac{1}{2} \log \frac{2E}{m} 
\,.
\end{align}

Let us consider an energy \(E \approx \SI{e15}{eV}\). Then, \(y_0 \approx 7\). 

We can divide the CMS depending on the sign of \(\eta^{*}\),the pseudo-rapidity in the CMS. Then the total rapidity will read \(\eta = \eta^{*} + y_0 \). 

The boundary between these in the lab frame is found when \(\eta^{*} = 0\), so we get \(\eta = - \log \tan \theta /2\): \(\theta \) is about 6 arcminutes for \(\eta = 7\). 

For a collision \SI{20}{km} in the air, we expect about half of the particles to be found within a \SI{40}{m} radius. 

Collider experiments typically cover \(\abs{\eta } \lesssim 5\) (the figure refers to ATLAS);
this only covers the central region, we do not get all of them. 

The definition of the pseudo-rapidity is convenient because PDFs in the form \(\dv*{N}{\eta }\) are only shifted under Lorentz boosts. 

The \textbf{transverse mass} can also be defined: 
%
\begin{align}
E^2 = m^2 + p^2 = \underbrace{m^2 + p_T^2}_{m_T^2} + p^2_\parallel
\,,
\end{align}
%
and \(m_T^2 = m^2 + p_T^2\) defined as above is Lorentz invariant under boosts in the \(\parallel\) direction. 
We can also write this relation like 
%
\begin{align}
\qty( \frac{E}{m_T})^2 - \qty( \frac{p_\parallel}{m_T})^2 &= 1  \\
\cosh^2 y - \sinh^2 y &= 1  
\,.
\end{align}

The rapidity can also be expressed as 
%
\begin{align}
y = \frac{1}{2} \log \qty(\frac{E + p_\parallel}{m_T})
\,.
\end{align}

Exercise: suppose we have an incoming particle with mass \(M\) impacting upon a particle with mass \(m\); what is the maximum transfer of energy from particle \(M\) to particle \(m\), if the collision is elastic? 

The result should be 
%
\begin{align}
\Delta E = \frac{2 m \gamma^2 \beta^2}{1 + 2 \gamma m/M + (m/M)^2}
\,.
\end{align}

The case we are interested in will be a proton interacting with a medium and interacting with an electron. There, and for a relativistic proton, this will read \(\Delta E \approx 2 m \gamma^2\). 

Next time, we will discuss some matter-radiation interactions. 

\begin{extracontent}
A particle of mass \(M\) with Lorentz factor \(\gamma \) and velocity \(\beta \) is impacting a stationary particle of mass \(m\). 

The total invariant mass will be 
%
\begin{align}
\sqrt{s} &= \sqrt{(E_M + m)^2  - p_M^2}  \\
&= \sqrt{M^2 + m^2 + 2 \gamma Mm}
\,,
\end{align}
%
where \(E_M = \gamma M\) while \(p_M = \beta E_M\).

The Lorentz factor \(\gamma _{\text{CoM}}\) of the transformation required to move between the center of mass frame and the lab one can be computed from the total energy: 
%
\begin{align}
\gamma _{\text{CoM}} = \frac{E_{\text{tot}}}{\sqrt{s}} = \frac{\gamma M + m}{\sqrt{M^2 + m^2 + 2 \gamma Mm}}
\,.
\end{align}

As one would expect, this approaches 1 as \(\gamma \to 1\). 

In the CoM frame we will have 
%
\begin{align}
\sqrt{s} = E^{*}_M + E^{*}_m = E^{* \prime}_M + E^{* \prime}_m
\,,
\end{align}
%
where the prime denotes a value computed after the collision. 

Let us denote the momentum of particle \(m\) after the collision in the center of mass frame as \(\vec{p}^{*}\), so that 
%
\begin{align}
p_M^{* \prime} &= [E^{* \prime}_M, -\vec{p}^{*}] \\
p_m^{* \prime} &= [E^{* \prime}_m, \vec{p}^{*}]
\,,
\end{align}
%
the sum of which will equal \([\sqrt{s}, 0]^{\top} = p^{*\prime}_M + p^{*\prime}_m\).
Isolating the \(M\) momentum and computing the modulus of this equation yields the following expression for the energy of \(m\) after the collision:
%
\begin{align}
E_m^{*\prime} &= \frac{s + m^2 - M^2}{2 \sqrt{s}}  \\
&= \frac{M^2 + m^2 + 2 \gamma Mm + m^2 - M^2}{2 \sqrt{M^2 + m^2 + 2 \gamma Mm}}  \\
&= \frac{m^2 + \gamma Mm}{\sqrt{M^2 + m^2 + 2 \gamma Mm}}
\,.
\end{align}

As expected, this equals \(m\) when \(\gamma = 1\). 
Also, this precisely equals \(m \gamma _{\text{CoM}}\): this is correct, since what we have done is boosted particle \(m\) from stationarity with \(\gamma _{\text{CoM}}\); the kinematics before and after the collision are exactly the same, so the energy of particle \(m\) in the center of mass frame will be the same before and after it. 

In the CoM frame this is fixed, but the angle of emission is arbitrary, and can lead to differing values for the energy of \(m\) in the lab frame.

Specifically, the expression for the energy and momentum along the boost direction of mass \(m\) in the lab frame reads 

\begin{align}
\left[\begin{array}{c}
E_m \\ 
p_{m \parallel}
\end{array}\right] = \left[\begin{array}{cc}
\gamma _{\text{CoM}} & - \beta _{\text{CoM}} \gamma _{\text{CoM}} \\ 
- \beta _{\text{CoM}} \gamma _{\text{CoM}} & \gamma _{\text{CoM}}
\end{array}\right]
\left[\begin{array}{c}
E^{*\prime}_m \\ 
p^{*}_\parallel
\end{array}\right]
\,,
\end{align}

where \(p_\parallel^{*} = p^{*} \cos \theta \) is the momentum along the boost direction for mass \(m\). 
The value of \(p^{*}\) is fixed: we know \(\gamma_m^* = E^{*}_m / m\), and the corresponding value for \(\beta _m^*\) is readily computed as \(\beta _m^* = \sqrt{ 1 - (1 / \gamma _m^*)^2}\). Then, we will have \(p^{*}_m = \beta_m^* \gamma _m^* m\). 

Further, we can simplify the expression by making use of the fact that \(\gamma _{\text{CoM}} = \gamma_{m}^{*}\), and similarly for \(\beta _{\text{CoM}}\). 
With all this, the energy \(E_m\) in the lab frame reads 
%
\begin{align}
E_m 
&= \gamma _{\text{CoM}} \gamma_m^* m - \gamma _{\text{CoM}} \beta _{\text{CoM}} \gamma_m^* \beta _m^* m \cos \theta  \\
&= \qty(1 - \beta _{\text{CoM}}^2 \cos \theta ) \gamma _{\text{CoM}}^2  m  \\
&= \qty(1 - \beta _{\text{CoM}}^2 \cos \theta )
\gamma^2 _{\text{CoM}}
m 
\,.
\end{align}

This will be maximized when \(\cos \theta \) is as small as possible, so we require \(\cos \theta = -1\);\footnote{While setting \(\theta = 0\) yields the minimum, \(E_m = m\): the collision is perfectly elastic and time-symmetric,  \(m\) is bounced straight back in the CoM frame, so it returns stationary in the lab frame.} this means we get 
%
\begin{align}
E_m &= \qty(1 + \beta^2 _{\text{CoM}}) 
\gamma^2 _{\text{CoM}}
% \frac{(\gamma M + m)^2}{M^2 + m^2 + 2 \gamma Mm}
m  \\
&= \qty(2 - \frac{1}{\gamma _{\text{CoM}}^2}) 
\gamma^2 _{\text{CoM}}
% \frac{(\gamma M + m)^2}{M^2 + m^2 + 2 \gamma Mm}
m  \\
&= \qty(2 \gamma^2 _{\text{CoM}} - 1) m  
\,.
\end{align}

Before proceeding, we should remember that the quantity we are after is not the final energy \(E_m\) but the \emph{change} in energy of particle \(m\), which starts out stationary, so \(\Delta E = E_m - m\):
%
\begin{align}
\Delta E &=  2 \qty(\gamma^2 _{\text{CoM}} - 1) m\\
&= 2\qty( \frac{(\gamma M + m)^2}{M^2 + m^2 + 2 \gamma Mm}
- 1) m \\
&= 2 \frac{\qty(\gamma^2 M^2 + m^2 + 2 \gamma Mm) - M^2 - m^2 - 2 \gamma M m}{M^2 + m^2 + 2 \gamma Mm} m  \\
&= 2 \frac{M^2 \qty(\gamma^2 - 1)}{M^2 + m^2 + 2 \gamma Mm} m  \\
&= 2 \frac{\gamma^2 \beta^2}{1 + 2 \gamma m/M + (m/M)^2} m
\,.
\end{align}
\end{extracontent}
\end{document}
