\documentclass[main.tex]{subfiles}
\begin{document}

\marginpar{Thursday\\ 2021-12-9}

Natural radioactivity background is also given by \(\alpha \) particles. 

Background can be peaky (photoelectric effect, \(\alpha \) in the bulk of the detector) --- 
\(\alpha \)s near the edge can go outside the detector and therefore not lose all the energy inside.

It can also be continuous: Compton scattering and such. 

If we cannot discriminate \(\alpha \)s from photons we have no hope to measure a signal. 

Depending on how our calorimeter works, we can get different signals for the same energy 
for \(\alpha \)s and photons. 

By carefully cleaning the surfaces for extraneous radioactive elements 
they are able to reach \SI{e-2}{counts/ KeV / kg / yr} in the relevant region. 

Ideally, we'd like to have \(n_B \lesssim 1/ (\Delta E M t)\). 

If we have \(M = \SI{100}{kg}\), \(t = \SI{1}{yr}\), a \(Q\)-value of \SI{3}{MeV}, 
an energy resolution of \SI{1}{\percent}, 
we get a background index of \(n_B = \SI{3.3e-4}{kg^{-1} yr^{-1} KeV^{-1}}\). 

There are several strategies: 
\begin{enumerate}
    \item try to get a lot of material;
    \item squeezing \(\Delta E\);
    \item tracking or tagging events to make the background zero.
\end{enumerate}

We can either do tracking or calorimetry, basically. 
The main disadvantage of the latter is the difficulty of background rejection. 

Calorimeters can be scintillation or ionization based for the gas or liquid case, 
crystals instead can be both scintillating or non-scintillating 
(when they measure temperature). 

NEMO at Frejus was a nice experiment with tracking. 

Making smaller diodes means we get better resolution. 

We can do pulse shape discrimination to veto signals from \(\alpha \)s, and only be 
left with electrons. 
GERDA does this, and has put a strong limit on \(T^{0 \nu }_{1/2} \gtrsim \SI{1.8e26}{yr}\). 

It has \(n_B \Delta E \sim 1 / \SI{500}{kg yr}\). 

You run your experiment until you have funding for the next stage. 
The enrichment is tough: you have to deposit a check for them to start doing it.



\end{document}
