\documentclass[main.tex]{subfiles}
\begin{document}

\section*{Fri Dec 06 2019}

From last time we have the metric 
%
\begin{align}
  \dd{s^2} = \eta_{\mu \nu } \dd{x^{\mu }} \dd{x^{\nu }}  - \frac{4GJ}{r} \sin^2 \theta \dd{t} \dd{\varphi }
\,,
\end{align}
%
and since in polar coordinates the \(z\) axis is singular we use cartesian ones: 
%
\begin{subequations}
\begin{align}
  x &= r \sin \theta \cos \varphi  \\
  y &= r \sin \theta \sin \varphi 
\,,
\end{align}
\end{subequations}
%
which means \(\varphi = \arctan (y/x)\). Then 
%
\begin{align}
  \dd{\varphi } = \frac{1}{1 + y^2 / x^2} \dd{\qty(y/x)}
  = \frac{x \dd{y} - y \dd{x}}{r^2 \sin^2\theta } \dd{\varphi }
\,,
\end{align}
%
so we get 
%
\begin{align}
  \dd{s^2} = \eta_{\mu \nu } \dd{x^{\mu }} \dd{x^{\nu }}  
  - \frac{4GJ}{r^2} \frac{x \dd{y} - y \dd{x}}{r} \dd{t} 
\,,
\end{align}
%
since the \(\sin^2 \theta\) simplified. 
So our metric is Minkowski plus a perturbation: 
%
\begin{align}
  \delta g_{01} = \delta g_{10} = \frac{2GJy}{(x^2+ y^2+z^2)^{3/2}}
\,,
\end{align}
%
and 
%
\begin{align}
    \delta g_{02} = \delta g_{20} = \frac{-2GJx}{(x^2+ y^2+z^2)^{3/2}}
  \,,
\end{align}
%
which means that our Christoffel symbols are 
%
\begin{align}
  \Gamma^{\mu }_{\alpha \beta } = \frac{1}{2} \eta^{\mu \gamma } \qty(\delta g_{\gamma \beta, \alpha } + \delta  g_{\gamma \alpha , \beta } - \delta g_{\alpha \beta , \gamma })
  + O(J^2)
\,.
\end{align}

The 4-velocity is \(u^{ \alpha } = (u^{t}, 0, 0, u^{z})\), and the spin vector initially is \(s^{\alpha } _{\text{in}} = (0, s^{x} _{\text{in}}, 0,0)\).

We want to show that \(s^{z}\) is zero at all times. 

The equation is 
%
\begin{align}
  \dv{s^{z}}{\tau } + \Gamma^{z}_{\alpha \beta } u^{\alpha } s^{\beta }= 0
\,,
\end{align}
%
but the velocity has only two components: then 
%
\begin{align}
  \dv{s^{z}}{\tau } + \Gamma^{3}_{0 \beta } u^{0} s^{\beta } + \Gamma^{3}_{3 \beta } u^{3} s^{\beta } = 0
\,,
\end{align}
%
but 
%
\begin{align}
  \Gamma^{3}_{0 \beta } = \frac{1}{2} \eta^{3 3 } \qty(
      \delta g_{3 \beta , 0 }
      + \delta g_{03, \beta }
      - \delta g_{0 \beta , 3}
  )
\,,
\end{align}
%
but the first two are zero since they contain an index \(3\). The last is also zero: 
%
\begin{align}
  \delta g_{0 \beta , 3} = \pdv{}{z} \qty(\frac{2GJy}{r^{3}}) \quad \text{or}\quad 
  \pdv{}{z} \qty(-\frac{2GJx}{r^{3}})
\,,
\end{align}
%
and in either case we need to calculate the derivative on the \(x=y=0\) axis, therefore since the terms are proportional to \(x\) or \(y\) they will vanish. 

For the other term we also have
%
\begin{align}
  \Gamma^{3}_{3 \beta } = \frac{1}{2} \eta^{33} \qty(\delta g_{3 \beta , 3} + \delta g_{33, \beta } - \delta g_{\beta 3, 3}) =0 
\,,
\end{align}
%
so \(s^{z} = 0\) for all times. 

Now let us write the evolution for the \(s^{t}\) component. 
%
\begin{align}
  \dv{s^{t}}{\tau } + \Gamma^{0}_{\alpha \beta } u^{\alpha } s^{\beta } = 0
\,,
\end{align}
%
and then we must look at the Christoffels: 
%
\begin{subequations}
\begin{align}
\Gamma^{0}_{0 \beta } &= \frac{1}{2} \eta^{00} \qty(
    \delta g_{0 0, \beta } +
    \delta g_{0 \beta , 0} 
    - \delta g_{0 \beta ,0 } 
) =0 \\
\Gamma^{0}_{3 \beta } &= \frac{1}{2} \eta^{00} \qty(
    \delta g_{0 \beta , 3 } +
    \delta g_{3 0 , \beta } 
    - \delta g_{3 \beta ,0 } 
) =0
\,,
\end{align}
\end{subequations}
%
so the only components which evolve are 
%
\begin{subequations}
\begin{align}
  \dv{s^{1}}{\tau } + \Gamma^{1}_{\alpha \beta } u^{\alpha } s^{\beta } &= 0 \\
  \dv{s^{2}}{\tau } + \Gamma^{2}_{\alpha \beta } u^{\alpha } s^{\beta } &= 0
\,,
\end{align}
\end{subequations}
%
and now we notice this: since we need to form \(\delta g_{01}\) or \(\delta g_{02}\), one of the lowe two indices in the Christoffels must be zero: but it cannot be \(\beta \) since \(s^{t}=0\): so we set \(\alpha =0\), and consider only \(\Gamma^{1}_{0 \beta } u^{0} s^{\beta }\) and \(\Gamma^{2}_{0 \beta } u^{0} s^{\beta }\). Also, \(\beta \) can only be \(1\) or \(2\). Then the equations become: 
%
\begin{subequations}
\begin{align}
    \dv{s^{1}}{\tau } + \Gamma^{1}_{0 1 } u^{0 } s^{1 } + \Gamma^{1}_{0 2 } u^{0 } s^{2 } &= 0 \\
  \dv{s^{2}}{\tau } + \Gamma^{2}_{0 1} u^{0 } s^{1}  + \Gamma^{2}_{0 2} u^{0 } s^{2} &= 0
\,,
\end{align}
\end{subequations}
%
and as yesterday we can use \(u^{0} = \dv*{t}{\tau }\) to get derivatives with respect to \(t\): we then get 
%
\begin{subequations}
    \begin{align}
        \dv{s^{1}}{t } + \Gamma^{1}_{0 1 }  s^{1 } + \Gamma^{1}_{0 2 }  s^{2 } &= 0 \\
      \dv{s^{2}}{t } + \Gamma^{2}_{0 1}  s^{1}  + \Gamma^{2}_{0 2}  s^{2} &= 0
    \,.
    \end{align}
\end{subequations}

Now we can compute the Christoffels: 
%
\begin{align}
  \Gamma^{1}_{01} = \frac{1}{2} \eta^{11} \qty(
      \delta g_{11, 0} + \delta g_{01, 1}
  ) = 0
\,,
\end{align}
%
and similarly for \(\Gamma^{2}_{02}=0\). 

On the other hand: 
%
\begin{subequations}
\begin{align}
  \Gamma^{1}_{02} &= \frac{1}{2} \eta^{11} \qty(
      \delta g_{12, 0} + \delta g_{01, 2} 
      - \delta g_{02, 1}
  ) \\
  \Gamma^{2}_{01} &= \frac{1}{2} \eta^{22} \qty(
    \delta g_{21, 0} + \delta g_{02, 1} 
    - \delta g_{01, 2}
) = - \Gamma^{1}_{02}
\,,
\end{align}
\end{subequations}
%
and the time derivative vanishes: so in the end we get 
%
\begin{align}
  \Gamma^{1}_{02} = \frac{1}{2} \qty(\pdv{}{y} \qty(\frac{2GJy}{(x^2+y^2+z^2)^{3/2}}) - \pdv{}{x} \qty(- \frac{2GJx}{(x^2+y^2+z^2)^{3/2}}))
\,,
\end{align}
%
which we must calculate at \(x=y=0\): therefore the only relevant contribution to the derivative is the one in which the derivative acts on the numerator, since otherwise we get a term proportional to either \(x\) or \(y\). So in the end we get: 
%
\begin{align}
  \Gamma^{1}_{02} = - \Gamma^{2}_{01} = \frac{2GJ}{z^3}
\,.
\end{align}

In the end then 
%
\begin{subequations}
\begin{align}
  \dv{}{t} \left[\begin{array}{c}
  s^{x} \\ 
  s^{y}
  \end{array}\right] =
  \frac{2GJ}{z^3}
  \left[\begin{array}{cc}
  0 & 1 \\ 
  -1 & 0
  \end{array}\right]
  \left[\begin{array}{c}
    s^{x} \\ 
    s^{y}
\end{array}\right]
\,,
\end{align}
\end{subequations}
%
which, if \(z\) were constant, would be an angular precession with angular velocity \(\Omega = 2GJ/z^3\), called the Lense-Thirring precession. 

\(z\) is not actually a constant: however this describes the instantaneous angular velocity. 

We assumed that the spin vector and the angular momentum of the BH are orthogonal. In general, one finds the following expression in terms of the angular momentum vector \(\vec{J}\) and of the radial unit vector \(\hat{e}_{r}\):
%
\begin{align}
  \vec{\Omega}_{LT} = \frac{GJ}{c^2r^3} \qty(3 (\vec{J} \cdot e^{\hat{r}}) e^{\hat{r}} - \vec{J})
\,,
\end{align}
%
which notably has the same factor as the electric field of a dipole. It reduces to our formula if \(\vec{J}\) is parallel to \(e^{\hat{r}}\). 

\subsection{The Kerr metric}

Now we treat Kerr geometry: from 1963, the vacuum solution of a rotating spherical mass. It is: 
%

%
\boxalign{
\begin{align}
\begin{split}    
\dd{s^2} = 
&- \qty(1 - \frac{2GMr}{\rho^2}) \dd{t^2} - 
\frac{4GMar \sin^2\theta }{\rho^2} \dd{t} \dd{\varphi }
+ \frac{\rho^2}{\Delta } \dd{r^2} \\
&+ \rho^2 \dd{\theta^2} + \qty(r^2+a^2 + \frac{2GMr a^2 \sin^2\theta }{\rho^2}) \sin^2 \theta \dd{\varphi^2}
\end{split}
\,,
\end{align}}
%
where \(a = J/M\), \(\rho^2= r^2+a^2 \cos^2\theta  \) and \(\Delta = r^2 - 2GMr +a^2\). 

Recall that in \(c=1\) velocities are dimensionless, so the units of angular momentum are mass times length, therefore the term \(a = J/M\) is a length. 

At \(O(a^{0})\) we recover the Schwarzschild metric. 

At \(O(a^{1})\) we have the slowly rotating geometry. 

At \(r \gg GM\) we have 
%
\begin{subequations}
\begin{align}
  \begin{split}    
  \dd{s^2} = &- \qty(1 - \frac{2GM}{r}) \dd{t^2}
  - \frac{4GMa}{r} \sin^2 \theta \dd{t} \dd{\varphi } \\
  &+ \qty(1 + \frac{2GM}{r}) \dd{r^2} 
  + r^2 \qty(\dd{\theta^2} + \sin^2 \theta \dd{\varphi })
  \end{split}
\,,
\end{align}
\end{subequations}
%
where we approximated \(r^2/(r^2-2GMr) \sim 1 + \frac{2GM}{r}\). 

Then, we can orbit far away with a gyroscope and measure \(M\) and \(J\). 

We still have the Killing vectors \(\xi^{\mu } = (1, \vec{0})\) and \(\xi^{\mu }= (0,0,0,1)\). 

Notice that we have the symmetry \(\theta \rightarrow \pi - \theta \). We have a real singularity at \(\rho =0\): this means \(r =0 \) and \(\cos \theta =0\). 

\(r=0\) is not a single point! This can be seen from the fact that to go from \((r, \theta ) = (0, \theta_0 )\) to \((0, \theta_1 )\) we have a positive distance since the \(\dd{\theta^2} \) metric element is nonzero if \(a^2 \cos^2\theta \neq 0\). 

We have a coordinate singularity, an horizon, when \(\Delta = 0\): this means 
%
\begin{align}
  r^2 - 2GMr + a^2 = 0 \implies r_{\pm} =GM \pm \sqrt{G^2M^2-a^2}
\,,
\end{align}
%
which means that we have an outer horizon and an inner horizon. As \(a=0\) we have \(r_+ = 2GM\), which is fine.
However if \(a> GM\) we have a \emph{naked singularity}: there is a postulate called the \emph{cosmic censorship} postulate which states that naked singularities do not exist. 
This is not a theorem: it is just something which is suggested by black hole formation. 

The case in which \(a = GM\) is called the extreme Kerr solution. 

We are going to focus our attention on \(r_{+}\), the largest horizon. 

First of all: \emph{$r_{+}$ is a null surface}. 

It is a boundary which separates the region where light can go to \(r \rightarrow \infty \) from the one in which light eventually always goes to \(r \rightarrow 0\). 

Then light is ``trapped inside this surface''.

The simplest case of a null surface  is a light cone in Minkowski: in polar coordinates, the set of vectors in the form \(x^{\mu } = (\alpha , \alpha , \beta , \gamma )\).

We can write this as 
%
\begin{align}
  x^{\mu } = \alpha l^{\mu } + \beta m^{\mu } + \gamma n^{\mu }
\,,
\end{align}
%
and then we can see that \(l^{\mu } l_{\mu } = \alpha^2 - \alpha^2 = 0\). \(m^{\mu } \) and \(n^{\mu }\) are instead spacelike vectors. 

The set \((l, m, n)\) is  a basis for the tangent space of the light cone. 
So, not all the vectors tangent to the surface are null. 

For the Schwarzschild horizon instead we can select 
%
\begin{subequations}
\begin{align}
  l^{\mu } &= (1, 0, 0, 0)  \\
  m^{\mu } &= (0,0,1, 0)  \\
  n^{\mu } &= (0,0,0, 1)
\,,
\end{align}
\end{subequations}
%
and as before \(m\) and \(n\) are spacelike, while 
%
\begin{align}
  l \cdot l = - \qty(1 - \frac{2GM}{r}) (l^{0})^2 = 0 
\,,
\end{align}
%
since we are precisely at \(r = 2GM\). 

At the Kerr horizon instead we have: 
%
\begin{align}
  l^{\alpha } = (1, 0, 0, \Omega_{H})
\,,
\end{align}
%
where  \(\Omega_{H}= a / (2GMr_+)\): light will rotate along the Kerr black hole, while in the Schwarzschild case the light is stationary with respect to the spatial coordinates. 

Now let us take a picture of the horizon at fixed \(t\): 
it does not matter which time by stationarity. It will be homework to show that in that case on the horizon the metric will be 
%
\begin{align}
\dd{s^2} = \rho_{+}^2 \dd{\theta^2} + \qty(\frac{2GMr_+}{\rho_{+}})^2 \sin^2 \theta \dd{\varphi^2}
\,.
\end{align}

This is not a spherical surface, since \(\rho_{+}\) depends on \(\theta \). 

It can be shown that it looks like an ellipsoid. The equator is larger than the corresponding equator for a sphere. To show this we will compute the length of the equator: \(\theta = \pi /2\) and \(\varphi \in [0, 2 \pi ]\), and also we will compute the length of a north-south-north path: \(\varphi = \const\) and \(\theta \in [0, \pi ]\) and then \(\theta \in [\pi , 0]\).

For the equator we get 
%
\begin{align}
  L _{\text{equator}} = \int_{0}^{2 \pi } \dd{\varphi } \sqrt{g_{33}}
\,,
\end{align}
%
calculated at \(\theta = \pi /2\). This becomes 
%
\begin{align}
  L _{\text{equator}} = \int_{0}^{2 \pi } \dd{\varphi } 2GM = 4 \pi GM
\,,
\end{align}
%
since the \(\sin \theta = 1\) while \(\cos \theta = 0\). The harder one is the length of the circle: it comes out to be 
%
\begin{align}
  L _{\text{NSN}} = 2 \int_{0}^{\pi }  \dd{\theta } \sqrt{g_{\theta \theta }}
  = 2 \int_{0}^{\pi } \dd{\theta } \sqrt{r^2 + a^2 \cos^2\theta }
\,,
\end{align}
%
which we will expand for small \(a\) since we do not like elliptic integrals. This becomes 
%
\begin{align}
  L _{\text{NSN}} \approx 2 \int_{0}^{\pi } \dd{\theta } \qty(2GM + \frac{a^2}{4GM} \qty(-2 + \cos^2 \theta )  + O(a^2))
\,,
\end{align}
%
and since we are integrating over a period of \(\cos^2\theta\) we can substitute its average of \(1/2\): then the result is 
%
\begin{align}
L _{\text{NSN}} \approx 4\pi GM - \frac{3a^2 \pi }{4GM}
\,,
\end{align}
%
so the length is less than the equatorial one. 

\begin{bluebox}
Complement: what does the region \(r=0\) look like? 

Inserting \(r=0\) into the Kerr metric we find 
%
\begin{align}
  \dd{s^2} = - \dd{t^2} + \cos^2\theta \dd{r^2}
  + a^2 \qty(\cos^2 \theta \dd{\theta^2} + \sin^2 \theta \dd{\varphi^2})
\,,
\end{align}
%
and we can do the change of variable \(\psi = \sin \theta \): then the line element becomes 
%
\begin{align}
  \dd{s^2} = - \dd{t^2} + \qty(1 - \psi^2) \dd{r^2} +a^2 \qty(\dd{\psi^2} + \psi^2 \dd{\varphi^2})
\,.
\end{align}

This is not a point anymore: it actually consists of two disjoint ``emispheres'', and to see what the geometry of these looks like we can notice that at constant \(\theta \) we have a circle of radius \(\psi =\sin \theta \).

How does this surface actually look like? We can notice that the metric \(\dd{s^2}_{(2)} = a^2 \qty(\dd{\psi^2}  + \psi^2 \dd{\varphi^2})\) is, in polar coordinates, the metric for the plane, which is intrinsically flat: we do not need to actually compute the Riemann tensor to already know that all of its components are zero. 

Then, we have a bounded surface, made up of circles, which is flat so it can be made by folding a piece of paper; also it is axisymmetric and symmetric with respect to \(\theta  \rightarrow - \theta \): the surface is made up of two cones joined at the equator.

In a paper\footnote{\url{https://arxiv.org/abs/1205.5848v6}} I found the result that the aperture angle of the cone is 
%
\begin{align}
  2 \arctan \frac{\abs{a}}{\abs{GM}}
\,,
\end{align}
%
but I have not figured out how to prove it myself yet. 

\end{bluebox}

\end{document}