\documentclass[main.tex]{subfiles}
\begin{document}

\section*{Thu Jan 09 2020}

In the last four hours we will finish the discussion on GWs. 

We solved the vacuum equations \(R_{\mu \nu }=0\) for a perturbed Minkowski spacetime \(g_{\mu \nu } = \eta_{\mu \nu } + h_{\mu \nu } \) for small \(h_{\mu \nu }\). 

We found that these are equivalent to 
%
\begin{align}
R_{\mu \nu }  = - \frac{1}{2} \square h_{\mu \nu } + \partial_{(\mu } \qty(h^{\lambda }_{\nu) , \lambda } - \frac{1}{2} \partial_{\nu) }h )  =0
\,,
\end{align}
%
which can become the wave equation \(\square h_{\mu \nu } = 0\) in the harmonic gauge 
%
\begin{align}
\partial_{\lambda } h^{\lambda }_{\nu } - \frac{1}{2} \partial_{\nu } h =0
\,.
\end{align}

Our gauge transformations are 
%
\begin{align}
h_{\mu \nu } \rightarrow h_{\mu \nu } - \partial_{(\nu  } \epsilon_{\mu )}
\,,
\end{align}
%
where \(x^{\mu } \rightarrow x^{\mu } + \epsilon^{\mu }(x)\)

The solutions to the wave equation are 
%
\begin{align}
h_{\mu \nu } = C_{\mu \nu } \exp(i k \cdot x)
\,,
\end{align}
%
where \(k^2=0\), since \(p^{\mu } = \hbar k^{\mu }\) and the wave travels at the speed of light \(p^2=0\). 

\begin{claim}
If \(h_{\mu \nu }\) satisfies the harmonic gauge and \(\epsilon_{\mu }\) satisfies \(\square \epsilon_{\mu } = 0\) then \(\widetilde{h}_{\mu \nu } =  h_{\mu \nu } - \partial_{(\mu } \epsilon_{\nu )}\) also satisfies the harmonic gauge. 
\end{claim}

\begin{proof}
The equation we want to show is true is 
%
\begin{align}
0 = \partial_{\lambda } \widetilde{h}^{\lambda }_{\mu } - \frac{1}{2} \partial_{\mu } \widetilde{h}
\,,
\end{align}
%
where \(\widetilde{h} = \widetilde{h}^{\mu }_{\mu }\). 
We can expand it into 
%
\begin{align}
0 &= \partial_{\lambda } \qty(\cancelto{}{h^{\lambda }_{\mu }} - \partial^{\lambda } \epsilon_{\mu } - \partial_{\mu } \epsilon^{\lambda }) - \frac{1}{2} \partial_{\mu } \qty(\cancelto{}{h} - 2 \partial^{\nu } \epsilon_{\nu } )  \marginnote{Harmonic gauge for \(h_{\mu \nu }\)}\\
&= - \square \epsilon_{\mu } - \partial_{\mu } \qty(\partial_{\nu } \epsilon^{\nu }) + \partial_{\mu } \qty(\partial^{\nu } \epsilon_{\nu }) = 0 \,. \marginnote{\(\square \epsilon_{\mu } = 0\) and commuting derivatives}
\end{align}
\end{proof}

Now, we want to impose 4 additional conditions that remove the residual freedom \(x^{\mu } \rightarrow x^{\mu } + \epsilon^{\mu }\) with \(\square \epsilon^{\mu } = 0\). 

This is a wave equation for \(\epsilon^{\mu }\), so the solutions will look like 
%
\begin{align}
\epsilon_{\mu } = \gamma_{\mu } \exp( i k \cdot
 x ) 
\,,
\end{align}
%
with \(\gamma_{\mu }\) constant and \(k^2=0\). 

Under this transformation, the coefficient \(C_{\mu \nu }\) changes into 
%
\begin{align}
\widetilde{C}_{\mu \nu } = C_{\mu \nu } - 2i k_{(\mu } \gamma_{\nu )} 
\,,
\end{align}
%
as can be seen by substitutin in our expression for \(\gamma_{\mu } \) into the transformation law for \(h_{\mu \nu }\). 
Do note that these correspond to the same physical fields. 

We further require \(\widetilde{C}_{00 } = \widetilde{C}_{0i}= 0 \). This fixes the residual gauge freedom completely. 

We need to show that from a generic \(C_{\mu \nu }\) we can find a \(\gamma_{\mu }\) such that this condition holds. 

We write out the conditions: 
%
\begin{align}
\widetilde{C}_{00} = C_{00} - 2 i k_0 \gamma_0 
\,,
\end{align}
%
so setting \(\widetilde{C}_{00} =0 \) fixes 
%
\begin{align}
\gamma_0 = \frac{C_{00} }{2ik_0 }
\,,
\end{align}
%
while  in order to set 
%
\begin{align}
0=\widetilde{C}_{0i} &= C_{0i} - ik_0 \gamma_{i} - i k_{i} \gamma_0  \\
&= C_{0i} - i k_0 \gamma_{i} - i k_{i} \frac{C_{00} }{2 i k_0 }  \\
\gamma_{i } &= \frac{1}{i k_0 } \qty(C_{0i} - \frac{k_{i} C_{00} }{2 k_0 })
\,.
\end{align}

With these choices for \(\gamma_{\mu }\) we can impose the desired condition.

There is no issue if \(k_0 = 0\), since that would correspond to a GW of zero frequency, which can be shown to be equivalent to flat spacetime. 

In the end, we need to solve the system 
%
\boxalign{
\begin{align}
\square h_{\mu \nu } &=0  \label{eq:gw-wave-equation} \\
\partial_{\lambda } h^{\lambda }_{\mu } - \frac{1}{2} \partial_{\mu } h&=0  \label{eq:gw-harmonic-gauge}\\
h_{0\mu } &=0 \label{eq:gw-residual-gauge}
\,,
\end{align}}
%
since \(C_{\mu \nu }= 0\) for some \(\mu \nu \) implies \(h_{\mu \nu }=0\) for those indices.

\emph{A priori}, we have 10 degrees of freedom for our metric \(h_{\mu \nu }\) since it is symmetric. Equation \eqref{eq:gw-harmonic-gauge} eliminates 4 of them, and equation \eqref{eq:gw-residual-gauge} eliminates 4 more.

So, in the end we have 2 independent degrees of freedom. 

Let us write the harmonic gauge condition explicitly: for \(\mu =0\) we get
%
\begin{align}
\partial_{\lambda } h^{\lambda }_{0} - \frac{1}{2} \partial_{0} h &= 0  \\
- \partial_{0} h_{00} + \partial_{i} h_{i0} - \frac{1}{2} \partial_{0} h &= 0
\,,
\end{align}
%
but the first two vanish if \(h_{0\mu } =0\): so we get \(\partial_{0} h = 0\): this, when it is written explicitly, is 
%
\begin{align}
\partial_{0} \qty(C^{\mu }_{\mu } \exp(i k \cdot x)) = 0
\,,
\end{align}
%
which means that \(C^{ \mu }_{\mu } = C = 0\): so the gravitational wave is traceless, this can be written as \(h_{ii}=0\) since \(h_{00} =0  \). 

So, for \(\mu = j\) we can write the harmonic gauge condition as 
%
\begin{align}
\partial_{\lambda } h^{\lambda }_{j} &= 0 \marginnote{\(h = 0\)}  \\
\partial_{i} h_{ij} &=0 \marginnote{\(\eta_{ij} = \delta_{ij}\) and \(h_{0j} = 0\)} 
\,.
\end{align}
%

This implies that the GW is \emph{transverse}: it oscillates only in the directions perpendicular to the motion.

Then, our system becomes  
%
\boxalign{
\begin{align}
\square h_{\mu \nu } &= 0  \\
h_{0\mu } &= 0  \\
\partial_{i} h_{ij} &= 0  \\
h_{ii} &= 0
\,.
\end{align}}
%

In terms of the matrix \(C_{\mu \nu }\) these conditions mean: 
%
\begin{align}
C_{00} = C_{0i} = C_{ii} = k_{i} C_{ij} = 0 
\,,
\end{align}
%
since the condition \(\partial_{i} h_{ij} =0\) means \(i k_{i} C_{ij} \exp(i k \cdot x ) = 0\). 

Without  loss of generality we can write a coordinate system in which our gravitational wave is travelling along  the \(z\) direction: therefore \(\vec{k} = [0,0,k]^{\top}\). 
Then, we get the conditions: \(C_{i 3}=0\), \(C_{11}+ C_{22} = 0\) and \(C_{0 \mu } = 0\). 

So we have two conditions: \(C_{12} = C_{21} \neq 0\) and the rest of the matrix is zero, and \(C_{11} = - C_{22} \neq 0\) and the rest of the matrix is zero. Then, the basis tensors are 
%
\begin{align}
e_{ij,+} = \left[\begin{array}{ccc}
1 & 0 & 0 \\ 
0 & -1 & 0 \\ 
0 & 0 & 0
\end{array}\right]  
\,,
\end{align}
%
and 
%
\begin{align}
e_{ij, \times } = \left[\begin{array}{ccc}
0 & 1 & 0 \\ 
1 & 0 & 0 \\ 
0 & 0 & 0
\end{array}\right] 
\,.
\end{align}
%

The Frobenius norm of these matrices is \(e_{ij,+} e_{ij, +} = e_{ij, \times} e_{ij, \times } = 2\), which can also be written as \(\tr (e_{+} e_{+}) = \tr (e_{ \times  } e_{ \times }) = 2\). Also, in this sense they are orthogonal: \(\tr (e_{+} e_{ \times }) = 0\). 

So, we can write \(\tr (e_{r} e_{s}) = 2 \delta_{rs}\), where  \(r\) and \(s\) are either \(+\) or \( \times \). 

Then, any GW \(h_{ij}\) can be written as 
%
\begin{align}
h_{ij} (x) = \qty(h_{+} e_{ij, +} + h_{ \times } e_{ij, \times }) \exp(i k \cdot x)
\,,
\end{align}
%
with \(k^2 = 0\). Do note that \(h_{+} \) and \(h_{ \times }\) are two independent complex numbers. 

The solutions we want have real \(h_{\mu \nu }\), so if \(h_{\mu \nu } \in \mathbb{C}\) is a solution then \(h_{\mu \nu }^{*}\) also is a solution. So, we are allowed to take the real part with \((h_{\mu \nu } + h_{\mu \nu }^{*}) / 2\). This will then give us 
%
\begin{align}
\Re (C_{ij} \exp(i k \cdot
 x ) ) = C_{ij}  \cos (i k \cdot x  + \varphi )
\,,
\end{align}
%
for some real \(C_{ij}\). So, in general we will have 
%
\begin{align}
h_{ij}^{k} (x) = \sum_{r = +, \times } h_{r} e_{ij, r} \cos(kt - kz + \varphi_{r} )
\,,
\end{align}
%
do note that we changed the sign of the argument of the cosine, which we can do since the cosine is even. 

Do also note that the phase can in principle depend on the polarization. 

We can then recover any general solution travelling along the \(z\) direction by having \(h_r\) depend on \(k\): then  we will have 
%
\begin{align}
h_{ij} = \int  \dd{k} \sum_{r = +, \times } h_{r} (k) e_{ij, r} \cos(kt - kz + \varphi_{r}(k) ) 
\,.
\end{align}

Also, note that the frequency of the gravitational waves can be found by \(f = k / 2\pi \); this can be seen from the periodicity of the waves. 

What do these polarizations physically entail? 

Recall that the distance along a certain direction, say \(x\), is given by the integral \(\int \dd{x} \sqrt{g_{xx}}\). 

So, if \(g_{xx}\) changes than the physical distance also changes. 

Let us consider the point \(z=0\), at \(t=0\), with zero phase \(\varphi \) for the polarization \(e_{ij, +}\): then the component \(g_{xx}\) has increased, since the cosine is equal to one, while the component \(g_{yy}\) has decreased

After half a wavelength, when the cosine is equal to \(-1\), this effect reverses. Then, \(g_{xx}\) decreases while \(g_{yy} \) increases. 

The wave is travelling along the \(z\) axis, but because of the transversality condition it stretches the \(x\) and \(y\) directions. 

What happens if we perform a \SI{45}{\degree} rotation? the spatial rotation matrix is 
%
\begin{align}
R = \frac{1}{\sqrt{2}} \left[\begin{array}{ccc}
1 & 1 & 0 \\ 
-1 & 1 & 0 \\ 
0 & 0 & \sqrt{2}
\end{array}\right]
\,,
\end{align}
%
so, for example, \([1,1,0]^{\top} \rightarrow [\sqrt{2},0,0]^{\top}\). 

The transformation law with this matrix is \(\widetilde{x}^{i} = R_{j}^{i} x^{j}\).

It is known that a rotation matrix satisfies \(R^{\top} R = \mathbb{1}\). 

The line element changes as 
%
\begin{align}
\dd{\ell^2} = \dd{x^{i}} g_{ij} \dd{x^{j}} = \dd{\widetilde{x}^{i} } \widetilde{g}_{ij} \dd{\widetilde{x}^{j}}
\,,
\end{align}
%
which means that \(g = R^{\top} \widetilde{g} R\), since \(\dd{\widetilde{x}} = R \dd{x}\). This is coherent with the tensor transformation law. Inverting it, we find \(\widetilde{g} = R g R^{\top}\). 

How does the perturbed metric change? 
%
\begin{align}
\mathbb{1} + \widetilde{h} &= R (\mathbb{1} + h) R^{\top}  \\
&= R \mathbb{1} R^{\top} + R h R^{\top}  \\
&= \mathbb{1} + R h R^{\top} 
\,,
\end{align}
%
so \(\widetilde{h} = R h R^{\top}\). 

Let us suppose that \(h = e_{ \times }\): what is then \(\widetilde{h}\)? It can be shown straightforwardly that 
%
\begin{align}
\widetilde{h} = R e_{ \times } R^{\top} = e_{+}
\,,
\end{align}
%
so the cross-polarization is equivalent to the plus-polarization, rotated by \SI{45}{\degree}. 
This is somewhat of an informal line of reasoning, since the basis polarization ``tensors'' are not actually tensors under generic transformation. However, the result is correct and the reasoning is illustrative so we stick with it.

\end{document}
