\documentclass[main.tex]{subfiles}
\begin{document}

\section*{Thu Oct 10 2019}

Last lecture we introduced tensors.

An example of those is the EM tensor \(F_{\mu \nu}\):

\begin{equation}
  F_{\mu \nu} = \left[\begin{array}{cccc}
  0 & E_{x}/c & E_y/c & E_z/c \\ 
  -E_x/c & 0 & -B_x & B_y \\ 
  -E_y/c & B_x & 0 & -B_z \\ 
  -E_z/c & -B_y & B_z & 0
  \end{array}\right]\,,
\end{equation}
%
which, it can be checked, transforms as a \((0,2)\) tensor. Also, we can define the current vector \(j^{\mu} = (c \rho, j^{i})\). Then, the Maxwell equations read:

\begin{equation}
  \partial_{\mu} F^{\mu \nu }= \mu_0 j^{\nu}
  \qquad \text{and} \qquad
  \partial_{[\mu} F_{\nu \rho]}\,.
\end{equation}

They are covariant!

\subsection{Particles in motion}

In Newtonian mechanics, the motion of a particle is described by a function of time \(x^{i} = x^{i}(t)\).

In special relativity, we introduce the concept of \emph{worldline} of a particle: the one-dimensional set of spacetime events in which the particle is found.
It must be parametrized with respect to some parameter \(\lambda\), such that \(x^{\mu} = x^{\mu}(\lambda)\). A preferred choice for \(\lambda\) is the proper time of the particle, \(\lambda = \tau\), which is defined by \( \dd{s^2} = - \dd{\tau^2}\). 

We then define the 4-velocity:
%
\begin{equation}
  u^\mu = \dv{x ^\mu}{\tau} \,.
\end{equation}

It is a tensor since it is the product of a scalar and a tensor.

Multiplying \(u^\mu u_{\mu}\) we always get \(-c^{2}\), since:

\begin{equation}
  u^{\mu} u_{\mu} = \frac{\dd{x^{\mu}} \dd{x_{\mu}}}{\dd{\tau^{2}}} = -c^2 \frac{\dd{s^{2}}}{\dd{s^2}}
\end{equation}


We can make the expression of the 4-velocity explicit using \(\dd{\tau} = \gamma \dd{t}\), which gives us \(u^{\mu} = (\gamma c, \gamma v^{i})\).
In the frame of the particle the particle is not moving, therefore \(u^{\mu} = (c, 0)\).   

The \emph{four-momentum} of a massive particle with mass \(m\) is defined as: 

\begin{equation}
  p^{\mu} = m u^{\mu} = (m \gamma c, m \gamma v^{i})\,.
\end{equation}

The component \(p^{0}\) is equal to \(mc\) at \(v=0\). What does it look like in the nonrelativistic approximation? we can expand it for small \(v/c\):

\begin{equation}
  \frac{mc}{\sqrt{1 - \frac{v^{2}}{c^{2}}}} \sim
  mc \qty(1 + \frac{v^{2}}{2 c^{2} } )
  = mc + \frac{1}{c} \frac{mv^{2}}{2}\,.
\end{equation}

We get the mass, plus a kinetic energy term: more explicitly, \(cp^0 = mc^{2} + \frac[i]{1}{2} m v^{2}\). 

We can rewrite Newton's first law in SR:

\begin{proposition}[Newton I]
A free particle moves with constant \(u^{\mu}\), or 

\begin{equation}
  \dv{u^{\mu}}{\tau} = 0 
\end{equation}
\end{proposition}

To express this in an easier way we introduce the 4-acceleration:

\begin{equation}
  a^{\mu} \defeq \dv{u^{\mu}}{\tau} = \dv[2]{x^{\mu}}{\tau}
\end{equation}

We now wish to introduce the concept of a path minimizing proper time. 
Recall Snell's law, which allows us to relate the angles of incidence of light when it passes between one medium to another, if they have different indices of refraction:

\begin{equation}
  n_1 \sin(\theta_1 ) = n_2 \sin(\theta_2 )
  \qquad \text{or} \qquad
  \frac{\sin(\theta_{2})}{\sin(\theta_{1})} = \frac{n_1}{n_2 } = \frac{v_2}{v_1 }\,.
\end{equation}

It can be shown that a beam of light following this law is equivalent to the beam minimizing the time it takes to move from a point in one medium to a point in the other.

Analogously, saying that a massive particle travels along the worldline which minimizes \(\tau\) is equivalent to Newton's first principle.

We want to perturb a generic  worldline \(x^{\mu}\) with some \(\dd{x^{\mu}}\), and consider the proper time functional \(\tau\) which gives the proper time of a generic trajectory: we impose

\begin{equation}
  \frac{\tau \qty[x^{\mu} + \varepsilon^\mu] - \tau \qty[x^{\mu}]}{\abs{\varepsilon ^\mu} } = \frac{\delta \tau}{\delta x^\mu} \overset{!}{=}  0\,,
\end{equation}
%
where a limit \(\abs{\varepsilon^{\mu}} \rightarrow 0 \) is implied, and only the linear terms are considered.

The proper time functional for paths between \(A\) and \(B\) is given \(\tau = \int _A^B \dd{\tau}\).  
We can rewrite it as:

\begin{equation}
  \tau = \int_A^B \dd{\tau} \frac{\dd{\tau^2}}{\dd{\tau^2}}
  = \int_A^B \dd{\tau} \frac{-\eta_{\mu \nu}\dd{x^{\mu} \dd{x^{\nu}}}}{\dd{\tau^2}}\,.
\end{equation}

We now consider a perturbation \(\varepsilon^{\mu} = \delta^{\mu}_1 \delta x\):

\begin{equation}
  \tau_{AB}[x + \varepsilon] = \int_A^B \dd{\tau}
  \qty[\qty(\dv{t}{\tau})^2 
  - \frac{1}{c^{2}} \qty(\dv{t}{\tau} + \dv{\delta x}{\tau})^2 
  - \frac{1}{c^{2}} \qty(\dv{y}{\tau})^2
  - \frac{1}{c^{2}} \qty(\dv{z}{\tau})^2]\,.
\end{equation}

We can discard a second order term \(\qty(\dv*{\delta x}{\tau})^2\), and subtract off \(\tau_{AB}[x]\): we are left with 

\begin{equation}
  \delta \tau = - \frac{2}{c^{2}} \int_A^B \dd{\tau} \dv{x}{\tau} \dv{\delta x}{\tau} 
\end{equation}

Now, we integrate by parts, disregard the boundary terms since the endpoints of the path cannot be deformed, and get:

\begin{equation}
  \frac{\delta \tau_{AB}}{\delta x} = + \frac{2}{c^{2}} \int_A^B \dd{\tau} \dv[2]{x}{\tau} \,,
\end{equation}

which proves the equivalence for this type of perturbation, the others are analogous.

\begin{bluebox}
Let us write a more general sequence of equations, somewhat informally: 
%
\begin{subequations}
\begin{align}
\delta \qty(\int_{A}^{B} \dd{\tau }) &= 
- \delta \int  \frac{ \dd{x^{\mu }} \dd{x_{\mu }}}{ \dd{\tau^2}} \dd{\tau^2}  \marginnote{Expand \(\dd{s^2}\)} \\
&= - \int \frac{\qty( \dd{x_{\mu }} + \dd{\epsilon_{\mu }}) \qty( \dd{x^{\mu }} + \dd{\epsilon^{\mu }})}{ \dd{\tau^2}} \dd{\tau }  \marginnote{Perturb}\\
&= - 2 \int \frac{ \dd{\epsilon_{\mu } \dd{x^{\mu }}}}{ \dd{\tau^2}} \dd{\tau }  \marginnote{Discard terms of order different than 1 in the perturbation}\\
&= + 2 \int \epsilon_{\mu } \dv[2]{x^{\mu }}{\tau } \dd{\tau }
\overset{!}{=} 0
\marginnote{Integrate by parts}
\,,
\end{align}
\end{subequations}
%
which must hold for any \(\epsilon_{\mu }\): so, by the fundamental lemma of variational calculus, we must have 
%
\begin{align}
\dv[2]{x^{\mu }}{\tau } = 0
\,.
\end{align}
\end{bluebox}

The generalization of Newton's second law, which at low speeds is \(F^{i} = m a^{i}\), can be similarly restated as \(\delta S = 0\), for the action \(S = \int L \dd{\tau}\), where \(L\) is the Lagrangian for the system. 

\subsection{Motion of light rays}

For light we cannot compute \(u^{\mu}\) with the previous definition, since its proper time is always zero.

Instead, we \emph{define} \(u^{\mu}\) to be a normalized null-like vector, such that locally \(x^{\mu} = \lambda u^{\mu}\) for some \(\lambda\).

We know from quantum mechanics that \(E = \hbar \omega\), where \(\hbar = h / (2 \pi)\) and \(\omega = 2 \pi / T = 2 \pi f\). 

The momentum is proportional to the wavevector \(k^{i}\): \(p^{i} = \hbar k^{i} / c\). The relativistic generalization of this fact is 

\begin{equation}
  p^{\mu} = \qty(\frac{\hbar \omega}{c}, \frac{\hbar k^{i}}{c}) = \frac{\hbar k^{\mu}}{c}\,.
\end{equation}

Since the momentum of light must be null-like (\(p_{\mu } p^{\mu } = 0\)) we have that necessarily \(\omega = \abs{k} \).

\subsection{Doppler effect}

We take a special case: radiation goes in the same direction as the observer.
In the \(O\) frame we have \(k^{\mu} = (\omega, \omega,0,0)\).

The observer, moving with velocity \(v\), measures \(k^{\prime \mu}\). This can be easily computed with a Lorentz transformation: \(k^{\prime \mu} = \tensor{\Lambda}{^\mu_\nu} k^\nu\).

We are mostly interested in \(k^{\prime 0} = \omega'\): it comes out to be \(\omega' = \gamma \omega + (- \gamma \beta)\omega = (1- v/c) \gamma \omega\).

Some notes: at slow speeds \(\omega' \approx (1-v/c) \omega\); we have \(f'<f\) when source and observer are moving away from each other. This can be readily proven by considering either \(k^{\mu } = (\omega , -\omega , 0,0)\) or \(\beta \rightarrow - \beta \). 

\end{document}
